### 2017-01-23 Main.tex
First version

\documentclass[sigplan, anonymous, review]{acmart}
%\documentclass[prodmode,acmtoplas]{acmsmall}

\usepackage{color}
\usepackage{graphicx}
\usepackage{latexsym}
\usepackage{stmaryrd}
\usepackage{amsmath}
\usepackage{amssymb}
% \usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage{mathpartir}
\usepackage{wasysym}
\usepackage{url}
\usepackage{algorithm, algpseudocode}
\usepackage{fancyvrb}
%\usepackage[[style=1]{mdframed}
\usepackage{comment}
\usepackage{soul}

\definecolor{light}{gray}{.75}

% \mdfdefinestyle{MyFrame}{%
%     linecolor=white,
% %%    outerlinewidth=2pt,
% %%    roundcorner=20pt,
% %%    innertopmargin=\baselineskip,
% %%    innerbottommargin=\baselineskip,
% %%    innerrightmargin=20pt,
% %%    innerleftmargin=20pt,
%     backgroundcolor=light}

\newcommand{\gbox}[1]{\colorbox{light}{$\!\!#1\!\!$}}

%%\bibliographystyle{plain}

%% for comments
\input{macro-comments}
\input{macros}
\input{Preamble}
\input{lemmas}

% \newenvironment{example}{\par\noindent\textit{Example}\quad}{}

% Metadata Information
%\acmVolume{1}
%\acmNumber{1}
%\acmArticle{1}
%\acmYear{2012}
%\acmMonth{1}

% Document starts
\begin{document}

% Page heads
\markboth{B. Oliveira et al.}{The Implicit Calculus: A New Foundation for Generic Programming}

% Title portion
\title{The Implicit Calculus: A New Foundation for Generic Programming}
\author{Ommitted for anonymous submission}

\begin{abstract}
\input{src/Abstract}
\end{abstract}

\begin{comment}
\category{D.3.2}{Programming Languages}   
                {Language Classifications}
                [Functional Languages, Object-Oriented Languages]
\category{F.3.3}{Logics and Meanings of Programs}   
                {Studies of Program Constructs}
                []

\terms{Languages}

\keywords{
Implicit parameters, type classes, C++ concepts, generic programming,
Haskell, Scala}

\acmformat{Oliveira, B. C. d. S., Schrijvers, T., Choi, W., Lee, W., Yi, K., Wadler, P.
201?. The implicit calculus: a new foundation for generic programming.}

\begin{bottomstuff}
Author's addresses: 
B. C. d. S. Oliveira, Department of Computer Science, Hong Kong University; 
T. Schrijvers, Department of Applied Mathematics and Computer Science, Ghent University;
W. Choi {and} W. Lee {and} K. Yi, \ldots, Seoul National University;
P. Wadler, \ldots, University of Edinburgh.
\end{bottomstuff}
\end{comment}

\maketitle

\input{src/Introduction}

\input{src/Overview}

\input{src/Types}

\input{src/Translation}

\input{src/SourceLang}

%\input{src/Related}

\input{src/Conclusion}

% Appendix
\appendix
\section*{APPENDIX}
\setcounter{section}{1}

%\appendixhead{OLIVEIRA}

% Acknowledgments
\begin{acks}
We are grateful to Ben Delaware, Derek Dreyer, Jeremy Gibbons, Scott
Kilpatrick, eta Ziliani, the members of ROPAS and the
anonymous reviewers for their comments and suggestions.  This work was
partially supported by Korea Ministry of Education, Science and
Technology/Korea Science and Enginering Foundation's ERC grant
R11-2008-007-01002-0, Brain Korea 21, Mid-career Research Program 2010-0022061, and
by Singapore Ministry of Education research grant MOE2010-T2-2-073.
\end{acks}

% Bibliography
\bibliographystyle{ACM-Reference-Format}
\bibliography{sigproc}

%\input{src/Proofs}

% History dates
%\received{February 2007}{March 2009}{June 2009}

\end{document}

### 2017-01-23 src/Abs.tex
First version

Implicit programming that infers values by using type
information has proven its benefits in practice. However, 
years of experience have exposed several limitations in its current practice.
For example, Haskell's type classes, which are the oldest and most
prominent form of implicit programming have several
limitations of its original design: no
ability to control the \emph{scoping} of instances; the \emph{second
  class} nature of types classes compared to types; and type class
constraints being limited to \emph{first-order predicates}.

This paper presents a \textit{implicit calculus}: a powerful and expressive,
yet minimalistic, calculus that provides an implicit programming model
without the limitations of the current practice. The implicit calculus
offers control over scoping, does not suffer from the second class
nature of type classes, and incoorporates two features which
have been discussed in the literature but for which there is
currently no good solution: \emph{higher-order predicates} and
\emph{coherent support for overlapping instances}.  We present the
syntax, operational semantics, type system, type-directed translation
into System F, as well as correctness results. Finally, we show how
Haskell and Scala's implicit programming constructs are translated
into our calculus.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "../Main"
%%% End:

### 2017-01-23 src/Conclusion.tex
First version

\section{Conclusion}
\label{sec:conclusion}

Our main contribution is the development of the implicit
calculus $\ourlang$. This calculus isolates and formalizes the key
ideas of Scala implicits and provides a simple model for language designers 
interested in developing similar mechanisms for their own languages. 
In addition, $\ourlang$ supports higher-order rules and partial resolution, 
which add considerable expressiveness to the calculus.

Implicits provide an interesting alternative to conventional GP 
mechanisms like type classes or concepts. By decoupling resolution 
from a particular type of interfaces, implicits make resolution 
more powerful and general. Furthermore, this decoupling has other benefits too. 
For example, by modeling concept interfaces as conventional types, those interfaces can 
be abstracted as any other types, avoiding the issue of second class interfaces 
that arise with type classes or concepts.

Ultimately, all the expressiveness offered by $\ourlang$
offers a wide-range of possibilities for new generic programming applications.

%%In particular, 
%%resolution for any types local and nested scoping for implicits

%%features for
%%generic programming, nested scoping and resolution of any types.

%%We present a formal type system for the calculus and provide semantics via a
%%translation to System F. We prove that the translation preserves types and thus
%%establish type soundness for $\ourlang$.

%%The calculus provides a formal platform for the development of
%%a realistic source language. Our small source language already shows 
%%how to add implicit instantiation on top of the calculus. In further
%%work we intend to develop this into a full-fledged language.

% integrates key features in GP, while at the same
% time being minimalistic. The calculus supports nested scoping,
% overlapping rules, higher-order rules and rule resolutions. 
%
% and encodings of existing GP practice. 
% 
% In the calculus, two key features -- scoping and resolution -- of many
% GP mechanisms were put in the spotlight, and we
% offered answers to two challenging, and essentially unsolved issues in
% the literature: how to support coherence for overlapping rules in
% nested scoping; and how to support resolution with higher-order
% rules. Such properties of scoping and resolution have not received
% appropriate attention in the past. 
% 
% of GP, which has been an
% increasingly popular trend in programming languages such as Haskell, Scala,
% Java, C++ and Agda.

%%Both of these shortcommings hinder the development of applications 
%%that rely on sophisticated implicit programming mechanisms, as
%%document previously in the literature~\cite{derivable,}.

%Future work includes the design of a source language with a powerful 
%modularity construct (which could be a form of modules, OO classes 
%or records with associated types) to model ``type-class'' style
%interfaces. Type inference and additional source-level constructs on top of
%$\ourlang$ also deserve further attention.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "../Main"
%%% End:

### 2017-01-23 src/Over.tex
First version

%% ODER: format ==         = "\mathrel{==}"
%% ODER: format /=         = "\neq "
%
%
\makeatletter
\@ifundefined{lhs2tex.lhs2tex.sty.read}%
  {\@namedef{lhs2tex.lhs2tex.sty.read}{}%
   \newcommand\SkipToFmtEnd{}%
   \newcommand\EndFmtInput{}%
   \long\def\SkipToFmtEnd#1\EndFmtInput{}%
  }\SkipToFmtEnd

\newcommand\ReadOnlyOnce[1]{\@ifundefined{#1}{\@namedef{#1}{}}\SkipToFmtEnd}
\usepackage{amstext}
\usepackage{amssymb}
\usepackage{stmaryrd}
\DeclareFontFamily{OT1}{cmtex}{}
\DeclareFontShape{OT1}{cmtex}{m}{n}
  {<5><6><7><8>cmtex8
   <9>cmtex9
   <10><10.95><12><14.4><17.28><20.74><24.88>cmtex10}{}
\DeclareFontShape{OT1}{cmtex}{m}{it}
  {<-> ssub * cmtt/m/it}{}
\newcommand{\texfamily}{\fontfamily{cmtex}\selectfont}
\DeclareFontShape{OT1}{cmtt}{bx}{n}
  {<5><6><7><8>cmtt8
   <9>cmbtt9
   <10><10.95><12><14.4><17.28><20.74><24.88>cmbtt10}{}
\DeclareFontShape{OT1}{cmtex}{bx}{n}
  {<-> ssub * cmtt/bx/n}{}
\newcommand{\tex}[1]{\text{\texfamily#1}}	% NEU

\newcommand{\Sp}{\hskip.33334em\relax}

\newcommand{\Conid}[1]{\mathit{#1}}
\newcommand{\Varid}[1]{\mathit{#1}}
\newcommand{\anonymous}{\kern0.06em \vbox{\hrule\@width.5em}}
\newcommand{\plus}{\mathbin{+\!\!\!+}}
\newcommand{\bind}{\mathbin{>\!\!\!>\mkern-6.7mu=}}
\newcommand{\rbind}{\mathbin{=\mkern-6.7mu<\!\!\!<}}% suggested by Neil Mitchell
\newcommand{\sequ}{\mathbin{>\!\!\!>}}
\renewcommand{\leq}{\leqslant}
\renewcommand{\geq}{\geqslant}
\usepackage{polytable}

%mathindent has to be defined
\@ifundefined{mathindent}%
  {\newdimen\mathindent\mathindent\leftmargini}%
  {}%

\def\resethooks{%
  \global\let\SaveRestoreHook\empty
  \global\let\ColumnHook\empty}
\newcommand*{\savecolumns}[1][default]%
  {\g@addto@macro\SaveRestoreHook{\savecolumns[#1]}}
\newcommand*{\restorecolumns}[1][default]%
  {\g@addto@macro\SaveRestoreHook{\restorecolumns[#1]}}
\newcommand*{\aligncolumn}[2]%
  {\g@addto@macro\ColumnHook{\column{#1}{#2}}}

\resethooks

\newcommand{\onelinecommentchars}{\quad-{}- }
\newcommand{\commentbeginchars}{\enskip\{-}
\newcommand{\commentendchars}{-\}\enskip}

\newcommand{\visiblecomments}{%
  \let\onelinecomment=\onelinecommentchars
  \let\commentbegin=\commentbeginchars
  \let\commentend=\commentendchars}

\newcommand{\invisiblecomments}{%
  \let\onelinecomment=\empty
  \let\commentbegin=\empty
  \let\commentend=\empty}

\visiblecomments

\newlength{\blanklineskip}
\setlength{\blanklineskip}{0.66084ex}

\newcommand{\hsindent}[1]{\quad}% default is fixed indentation
\let\hspre\empty
\let\hspost\empty
\newcommand{\NB}{\textbf{NB}}
\newcommand{\Todo}[1]{$\langle$\textbf{To do:}~#1$\rangle$}

\EndFmtInput
\makeatother
%

\section{Overview} %%of Rule Calculus $\ourlang$}
\label{sec:overview}

In this section, we motivate and give an overview of the rule calculus and the
source language described in Section~\cite{}.

%%The rule calculus is
%%presented as a simple logic programming language of Horn clauses
%%and the relation to existing implicit programming mechanisms is
%%discussed. Then it is shown how a source language built on top 
%%of the rule calculus can model

\subsection{Rule Calculus $\ourlang$ Syntax}

\begin{figure}

\begin{center}

  \[ \begin{array}{llrl}
    \text{Types} & \type  & ::=  & \alpha \mid \rulet \\
    \text{Rule Types} & \rulet  & ::= & 
    \rulesch{\alpha}{\rulesetvar}{\type} \\
    \text{Context} & \rulesetvar  & ::= & \bar{\rulet} \\
    \text{Expressions} & e & ::=  & 
    ?\rulet \mid 
    \ruleabs{\rulet}{e} \mid
    e[\overline{\type}] \mid
    \ruleapp{e}{\rulesetexp} \\
  \end{array} \]
\end{center}

\caption{Rule calculus syntax}

\end{figure}

\subsection{Rules and Rule Types}

The key notion of the rule calculus is that of a rule and its
corresponding rule type. Rule types are

Relation to type classes; well-known relation between 
type classes and logic programming.

\paragraph{Horn clauses as rule types}

\Ex{
  1) \quad \{\} \To \tyint \\
  2) \quad \{\tybool\} \To \tyint \\
  3) \quad \{a \to a \to \tybool \} \To [a] \to [a] \to [a]  \\
  4) \quad \{\Eq~a\} \To [a] \to [a] \to [a] \\
  5) \quad \{\Eq~a, \Eq~b\} \To \Eq~(a,b)   
}

\paragraph{Polymorphic and higher-order rules}

\subsection{Resolution and Queries}

\paragraph{Higher-order queries}

\subsection{Scoping}

\paragraph{Local Scoping}

\paragraph{Nested Scoping}

\paragraph{Overlapping}

\subsection{Source Language and Type-Inference}

present $\ourlang$ and its core features using
some examples. The formal system of the calculus is presented in
Section~\ref{sec:calculus}. In the rest of this section, we assume
that the calculus is conservatively extended to have constructs of
$\lambda$-calculus and several types.

$\ourlang$ is a simple logic programming language of Horn clauses. We
can compose Horn clauses and resolve a clause under the given set of
clauses. Consider the following example:
\\
\Ex{
  \relation{f}{\myset{\tyint} \To \tyint} \\
  f = \ruleabs{(\myset{\tyint} \To \tyint)}{(?\tyint + 1)} \\
}
\\
In the example, $f$ is a {\it rule} whose type corresponds to Horn
clause $\tyint \To \tyint$. The clause, which we call {\it rule type},
means that $f$ will give us an integer value if there is an integer
clause in the assumptions. We call the current set of assumptions as
{\it rule environment}. In the body, the rule tries to resolve query
$?\tyint$ which essentially asks whether the rule type $\tyint$ can be
deduced from the given assumptions. Since $f$ assumes that there must
be a rule type $\tyint$, the query can be resolved. If we apply this
rule to a rule of type $\tyint$, we can get the final value like this:
\\
\Ex{ \ruleapp{f}{\myset{\relation{1}{\tyint}}} }
\\
This program will give us 2 as a result.

The rule type can be polymorphic in $\ourlang$. Consider the following
example:
\\
\Ex{
  \relation{g}{\myset{\forall \alpha. \alpha \to \alpha} \To 
    \tyint \to \tyint} \\
  g = \ruleabs{(\myset{\forall \alpha. \alpha \to \alpha} \To 
    \tyint \to \tyint)}{(?\tyint \to \tyint)} \\
}
\\
The query $?\tyint \to \tyint$ can be resolved by instantiating the
assumption $\forall \alpha. \alpha \to \alpha$ for the $\tyint$
type.

In the rest of the section, we explain the key features that
differentiate our calculus from the others. Note that some of these
features have already existed in the previous systems but none of them
provides the combination of all.

\subsection{Flexible Implicit Values}
\label{subsec:flexible}

We allow arbitrary types in rule types. This is different from type
class-like systems. The type class-like systems can be modeled as
simple logic language with Horn clauses of predicates on
types. Rather, our system is closer to Scala implicits~\cite{scala}.

Using arbitrary types for the assumptions has more flexibility than
using only predicates on types. Because of this limitation, some
Haskell programs try to encode natural number using type-level Church
numeral to do the very same thing like this:

\[  
\begin{array}{l}
  \relation{f}{\myset{\tyint} \To \tyint} \\ 
  f = \ruleabs{(\myset{\tyint} \To \tyint)}{(?\tyint + 1)}
\end{array}
\]
\[  
\begin{array}{l}
  \ruleapp{f}{\myset{\relation{\ruleabs{(\varnothing \To
          \tyint)}{1}}{\varnothing \To \tyint}}}
\end{array}
\]

\subsection{Lexical Scoping}
\label{subsec:lexical}

We support lexical scoping of rule environments. Consider the
following example:
\\
\[  
\begin{array}{l}
  \ruleapp
  {\ruleapp{f}{\myset{\relation{2}{\tyint}}}}
  {\myset{\relation{1}{\tyint}}}
\end{array}
\]
\\
The rule $f$ is now put under two environments.

Under the nesting, we choose the closest one from the given rule.

\subsection{Overlapping}
\label{subsec:overlapping}

We allow overlapping of rules.

\[
\begin{array}{l}
  \relation{f}{\forall \alpha. \alpha \to \alpha} \\
  f = 
  \ruleapp
  {(\ruleabs
    {(\forall \alpha.
      \myset{\forall \beta. \beta \to \beta, \tyint \to \tyint}
      \To \alpha \to \alpha)
    }
    {(?\alpha \to \alpha)})\\\quad\quad}
  {\myset{
      \relation{\lambda x. x + 1}{\tyint \to \tyint},
      \relation{\lambda x. x}{\forall \gamma. \gamma \to \gamma}
    }}
\end{array}
\]

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "../Main"
%%% End:

### 2017-01-23 src/Resolution.tex
First version

\section{Resolution with Overlapping Rules}
\label{sec:resolution}

An essential property of resolution is \textit{coherence}, which ensures that
the result of resolution is predictable and consistent. Incoherent programs,
where resolution cannot be resolved unambiguously, should be rejected.

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
\subsection{Overlap}
Overlapping rules are a well-known source of incoherence. Consider
\begin{equation*}
\left\{ \forall \alpha. \alpha \rightarrow \tyint,  
        \forall \alpha. \tyint \rightarrow \alpha \right\} \vturns \tyint \rightarrow \tyint
\end{equation*}
Both the first and the second rule can be used to discharge this judgement.
In general, we have incoherence when there are multiple possible resolution derivations.

In order to avoid incoherence, existing approaches in the literature take a the
Draconian measure of forbidding overlap altogether. However, overlap
is a highly useful features, as illustrated by the many practical
applications in GHC/Haskell.\footnote{despite the fact that Haskell
implementations do not truly rule out incoherence}
Hence, we adopt a less conservative approach.

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
\subsection{Overlap Resolution}
To avoid the potential incoherence, we provide a two-level priority scheme to
specify which one of several overlapping rules is selected during resolution.

The first priority is the natural priotirization in the implicit environment.
This environment behaves as a stack of sets. A rule in a set closer to the top
of the stack has priority over one further down in the stack. For instance,
\begin{equation*}
\left\{ \forall \alpha. \alpha \rightarrow \tyint \right\},  
\left\{ \forall \alpha. \tyint \rightarrow \alpha \right\} \vturns \tyint \rightarrow \tyint
\end{equation*}
is uniquely resolved using the second rule, which is at the top of the stack.

The second level, in case of
overlapping rules within the same set in the implicit environment, is to
select the most specific matching rule. For instance,
\begin{equation*}
\left\{ \forall \alpha. \alpha \rightarrow \alpha,  
        \forall \alpha. \alpha \rightarrow \tyint \right\} \vturns \tyint \rightarrow \tyint
\end{equation*}
is uniquely resolved using the second rule, which is more specific than the first.

While this approach allows us to make effective use of overlapping rules in many
cases, we must still avoid situations where no most specific rule can be determined,
such as the first example.
This overlap resolution also raises concerns with respect to type safety. Consider:
\begin{equation*}
\left\{ \forall \alpha. \alpha \rightarrow \alpha,  
        \{\textit{Char}\} \To  \tyint \rightarrow \tyint \right\} \vturns \beta \rightarrow \beta
\end{equation*}
Regardless of whether $\beta$ is \tyint or not, there is a rule to select
in the implicit environment. However, when $\beta$ is \tyint, resolution
gets stuck, because there is no rule to produce a value of type \texttt{Char}.
In contrast, resolution does not get stuck for other instantiations.
As we wish to claim that well-typed programs do not get stuck, we need to make
sure that situations like this do not happen.

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
\subsection{Condition Specifications}

In order to avoid unsafe and incoherent overlap in a ruleset, we impose three
conditions.

%~~~
\paragraph{Most Specific Rule}
Firstly, there must always be a most specific rule.
\begin{definition}[Existence of Most Specific Rule Condition]
We say that a ruleset $\pi$ always has a most specific rule iff
\[
\begin{array}{l}
  \forall \theta.\forall \tau.\rho_1,\rho_2\in\pi.
  \theta|\rho_1| \succeq \theta\tau \wedge \theta|\rho_2| \succeq \theta\tau \\
  \implies \exists \rho \in \pi. \theta|\rho| = \theta(|\rho_1| \sqcap |\rho_2|)
\end{array}
\]
We use $\mathit{exists}(\pi)$ to denote such ruleset.
\end{definition}

Note that the order between types can be affected by type type substitution.
A pair of types on which no order relationship exists may have an order after 
applying substiution. So we have to consider every possible substitutions.
See following example.
\[
\{
  \forall \alpha. \alpha \rightarrow {\tt int} \rightarrow a,
  \forall \alpha. {\tt int} \rightarrow \alpha  \rightarrow {\tt int} 
\} 
\vturns {\tt int} \rightarrow {\tt int} \rightarrow {\tt int}
\]

The resolution will be ambiguous if type variable $a$ is substituted with ${\tt int}$.

%~~~
\paragraph{Uniqueness}
Secondly, the most specific instance must be unique. This avoids situations
where multiple rules \textit{statically} have the same result type.
\begin{equation*}
\left\{ \texttt{int}, \texttt{char} \Rightarrow \texttt{int}, \texttt{char} \right\} 
\vturns \texttt{int}
\end{equation*}
However, we should also consider the effect of a type substitution, and make
sure that uniqueness remains true \textit{dynamically}. Consider this case:
\begin{equation*}
\left\{ \alpha, \tyint \right\} \vturns \tyint
\end{equation*}
It statically appears that the second is the only appropriate one to resolve
\tyint. However, at runtime, the type variable $\alpha$ may well be
instantiated to \tyint, which would mean that there are actually two
implicit \tyint values.

The following condition makes sure of both static and dynamic uniqueness.
\begin{definition}[Uniqueness of Instances]
We say that the ruleset $\pi$ is consistent with unique instances iff 
\[
\forall \rho_1,\rho_2 \in \pi, \theta. 
  \rho_1 \neq \rho_2
  \implies \theta |\rho_1| \neq \theta |\rho_2|
\]
We use $\mathit{unique}(\pi)$ to denote such ruleset.
\end{definition}

%~~~
\paragraph{Type Safety}
We must ensure that every resolution that succeeds for a general type also
succeeds for a more specific type.
\begin{definition}[Type Safety Condition]
An implicit environment $\env$ is type safe iff
\begin{equation*}
\forall \rulet, \theta  . \env \vturns \rulet \wedge \mathit{dom}(\theta) = \mathit{fv}(\rulet) \cup \mathit{fv}(\env)
~~\Longrightarrow~~ 
\theta\env \vturns \theta\rulet
\end{equation*}
\end{definition}

We may need restricted algorithmic check. Naively restricting specific instances to
have less constraint then general instances will not work. Assume we have the most
general pretty printer that prints $``\dots''$ for any input and polymorphic pretty
printer for lists which takes a pretty printer for an elemetns type implicitly.
A program having such pretty printers is natural but it will be rejected by naive
restriction

\tom{Algorithm}

%~~~
\paragraph{Coherent Environment} We now combine the above conditions
on rulesets and extend them to environments.
\begin{theorem}[Coherent Environment]
An environment $\env$ has a unique most specific instance for any query if
every stack element of the environment has unique most specific instances.
\tom{formal statement}
\end{theorem}
\wt{PROOF}

%-------------------------------------------------------------------------------
\subsection{Static Condition Checking}

We have formulated coherence conditions on the implicit environment, but not
yet stated where in the type system these conditions are imposed.

One approach is to impose the conditions throughout, on every 
environment and ruleset in all type system rules. This approach not only burdens the
formulation of the type system, it also unnecessarily rejects many useful
programs, like the following rule\footnote{similar to 
an \texttt{instance (C a, C b) => C (a,b)} of a type class \texttt{C} in Haskell}
\begin{equation*}
f \stackrel{\mathit{def}}{\equiv}\ruleabs{(\forall a,b.\{a,b\} \Rightarrow a \times b)}{(?a,?b)}
\end{equation*}
where $\{a,b\}$ violates the uniqueness condition (as $\theta a = \theta b$ for $\theta = [a \mapsto \type, b \mapsto \type]$). It would be a shame if we could
not use this rule in safe situations, like
\begin{equation*}
\ruleapp{f[\tyint,\texttt{Bool}]}{\{1:\tyint,\texttt{true}:\texttt{Bool}\}}
\end{equation*}
In the same vain, we would like to accept
\begin{equation*}
\ruleapp{f[\tyint,\tyint]}{\{1:\tyint\}}
\end{equation*}
where the substitution collapses the set of required implicit rules to a
singleton.
In order to avoid rejecting these programs, we must not check for uniqueness
at rule abstraction sites (\textbf{rule}), but defer them to a \textit{later} point.
In particular, we do want to reject programs like
\begin{equation*}
g \stackrel{\mathit{def}}{\equiv}
\ruleabs{(\forall a.\{a\} \Rightarrow a \times \tyint)}
      {\ruleapp
        {f[a,\tyint]}
        {\{?a:a,3:\tyint\}}}
\end{equation*}
which dynamically does not have a unique most general rule in a context like 
$\ruleapp{ g[\tyint]}{\{ 4 : \tyint\}}$. Deferring the check beyond
$g$'s definition leaves us with too little information. Hence, we must perform
the existence check at rule application (\textbf{with}).

Finally, we need another existence check at resolution ($?\rulet$). Consider
the program
\[\begin{array}{l}
\tstate \vturns 
\ruleapp{(?(\forall a.\{a\}\Rightarrow a \times \texttt{int}))[\texttt{int}]}
  {\{4:\texttt{int}\}}
\end{array}\]
where $\tstate = \{f:\forall a,b.\{a,b\}\Rightarrow (a, b) ,3:\texttt{int}\}$.
In this program, the two \texttt{int} values are supplied by different means to $f$,
$4$ through the \texttt{with} construct, but $3$ through recursive resolution. Hence,
the latter also requires a check.

\paragraph{Summary}
Our type system checks:  
\textit{existence} of most specific instances at rule definitions (\texttt{rule}), and
\textit{uniqueness} of instances at resolution ($?\rulet$) and rule application (\texttt{with}).
% If the ruleset in rule signatures statically ensures the existence of most specific instances, 
% then the runtime environment will always have most specific instances.
% Furthermore, if a static check prevents every attempt to make an environment with duplicate
% entries at rule application and sites, then we will always have unambiguous environments at runtime.
% There are only two such positions : rule applications and resolutions.

\begin{definition}[Coherent Program]
Program $e$ with type $\tau$ is coherent iff the following conditions hold throughout
the derivation tree of $\epsilon \turns e:\tau$.
\begin{itemize}
\item 
  For all subexpression $\ruleabs{\forall \overline{\alpha}.\pi\Rightarrow \tau}{e}$ :
    $\texttt{exists}(\pi)$
\item
  For all subexpression $\ruleapp{e}{\overline{e}:\overline{\rho}}$ :
    \mbox{$\texttt{unique}$}$(\overline{\rho})$
\item
  For all sub-derivations 
  $\Gamma \turns \forall \overline{\alpha}. \pi_1 \Rightarrow \tau$ :
    Assume $\Gamma(\forall \overline{\alpha}.\tau) 
    = \forall \overline{\beta}.\pi_2 \Rightarrow \tau_2$
    and $\theta$ is a substitution such that $\theta \tau_2 = \tau_1$.
    The the followings should hold : $\texttt{unique}(\theta\pi_2 - \pi_1)$
    and $\forall \rho_1 \in \pi_1.\rho_2 \in \pi_2. \theta\rho_2 \nsucceq \rho_1$.
\end{itemize}
\end{definition}

% Statically checking an argument at rule application sites is enough 
% to reject following ambiguous program. 
% \[\begin{array}{l}
% \texttt{let } f=\ruleabs{(\forall a,b.\{a,b\} \Rightarrow a \times b)}{(?a,?b)} \\
% \texttt{in let } g=
%   {\ruleabs{(\forall a.\{a\} \Rightarrow a \times \texttt{int})}
%     {\ruleapp
%       {f[a,\texttt{int}]}
%       {\{?a,3\}}}} \\
% \texttt{in } \ruleapp{g[\texttt{int}]}{\{4\}}
% \end{array}\]

% Static check at a resolution site can also prevent following ambiguous program.
% Let $I = \{(?a,?b):\forall a,b.\{a,b\}\Rightarrow a \times b  ,3:\texttt{int}\}$
% \[\begin{array}{l}
% I \vturns 
% \ruleapp{(\forall a.\{a\}\Rightarrow a \times \texttt{int})[\texttt{int}]}
%   {\{4:\texttt{int}\}}
% \end{array}\]

% Our checking algorithm can accept following case since we delay unambiguouty check to
% the use point.
% \[
% \ruleapp
%   {(\ruleabs{\forall a,b.\{a,b\} \Rightarrow (a,b)}{(?a,?b)})
%     [\texttt{int},\texttt{int}]}
%   {\{3:\texttt{int}\}}
% \]

\wt{THEOREM : Type Safe Program Never Generates Ambiguous Environment}
\tom{The theorem should just be type saftey.}
\wt{Mentioning other possible options}

%\begin{definition}[Most Specific Rule Condition]
%We say that an environment $\env$ always has a most specific rule iff
%\begin{multline*}
%\forall \tau.
%\forall \rho_1, 
%        \rho_2 \in \env^*(\tau).
%        \exists (\rulesch{\alpha_3}{\ruleset_3}{\tau_3}) \in \env^{*}(\tau).\\
%        \{\rho_1,\rho_2\} \subseteq \env^*(\tau_3)
%\end{multline*}
%where
%\begin{equation*}
%\left\{\begin{array}{rcl}
%(\env,\bar{\rho})^*(\tau) & = & \{(\rulesch{\alpha'}{\ruleset'}{\tau'}) \in \bar{\rho} \mid \tau \preceq_{\bar{\alpha}'} \tau' \} \lhd \env^*(\tau)\\
%\epsilon^*(\tau)          & = & \emptyset 
%\end{array}\right.
%\end{equation*}
%with
%\begin{equation*}
%\left\{\begin{array}{rcl}
%\emptyset \lhd \bar{\rho}_2      & = & \bar{\rho}_2\\
%\bar{\rho}_1 \lhd \bar{\rho}_2   & = & \bar{\rho}_1 \hspace{1cm}, \rho_1 \neq \emptyset
%\end{array}\right.
%\end{equation*}
%and
%begin{equation*}
%
%\tau_1 \preceq_{\bar{\alpha}} \tau_2 \stackrel{\mathit{def}}{\Longleftrightarrow}  
%\exists \bar{\tau}. [\bar{\alpha} \mapsto \bar{\tau}]\tau_2 = \tau_1
%\end{equation*}
%\end{definition}
%
%\tom{define lookup}
%
%
%The above condition only makes sure that every most step in the recursive
%resolution process either fails or succeeds deterministically.
%
%% and
%% \begin{multline*}
%% \forall (\rulesch{\alpha_1}{\kappa_1}{\ruleset_1}{\tau_1}), 
%%         (\rulesch{\alpha_2}{\kappa_2}{\ruleset_2}{\tau_2}) \in \ruleset.\\
%%         \tau_2 \preceq \tau_1 \wedge
%%         \tau_1 \preceq \tau_2 \Longrightarrow \\
%%         \rulesch{\alpha_1}{\kappa_1}{\ruleset_1}{\tau_1} = 
%%         \rulesch{\alpha_2}{\kappa_2}{\ruleset_2}{\tau_2}
%% \end{multline*}
%
%

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
\subsection{Operational versus Elaboration Semantics}
Our informal notion of coherent overlap always allows us to select
the appropriate implicit rule \textit{at runtime}, which is perfect
for the operational semantics.
Not so for the elaboration semantics, which (by definition) has to select the
appropriate rule statically. For instance,
\begin{equation*}
\left\{ \forall \alpha. \alpha \rightarrow \alpha,  
        \tyint \rightarrow \tyint \right\} \vturns \beta \rightarrow \beta
\end{equation*}
Statically, we do not know whether $\beta$ equals \tyint or not.  For the
operational semantics this is fine: statically we only check that there is a
unique resolution in both cases. At runtime we will know which of these to
chose, as the type variable $\beta$ will be instantiated.
However, the elaboration semantics must chose at compile time, and thus cannot
resolve this situation.

As a consequence, the elaboration semantics is forced to accept fewer programs
than the operational semantics.  In particular, if we insist that the implicit
environment is able to resolve every possible query coherently, then the
elaboration semantics cannot tolerate any overlap at all. A more liberal
approach only considers whether the queries $\env \vturns \rulet$ that actually
appear in the program (originating from the $?\rulet$ construct) are resolved
coherently. GHC takes a similar approach for overlapping type class instances,
yet coherence is not guaranteed across multiple modules. Note that only
checking coherence for program queries also makes sense for the operational
semantics.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "../Main"
%%% End:

### 2017-01-23 src/Termination.tex
First version

\section{Termination of Resolution}

If we are not careful about which rules are made implicit, the recursive
resolution process may not terminate. This section describes how to impose 
a set of modular syntactic restrictions that prevents non-termination. 
%% We have to avoid non-termination of the type checker to ensure soundness.

As an example of non-termination consider 
\begin{equation*}
\myset{
  \myset{\tychar} \To \tyint,
  \myset{\tyint} \To \tychar} \vturns \tyint
\end{equation*}
which loops, using alternatively the first and second rule in the implicit
environment.

The problem of non-termination has been widely studied in the context of
Haskell's type classes, and a set of modular syntactic restrictions
has been imposed on type class instances to avoid non-termination~\cite{fdchr}. 
Adapting these restrictions to our setting, we obtain the following termination
condition.

\defterm

% Haskell's condition is quite severe because the one global scope for all type
% class instances is \textit{open}: more instances can be added later (in other
% modules).  The modularity of the condition already anticipates such future
% additions.
% 
% In contrast, our local scopes are \textit{closed}. Later extensions of the
% program (e.g., new modules) do not affect the existing scopes. Hence, in
% $\ourlang$, termination of resolution coincides with the traditional program
% termination problem. So, alternatively, $\ourlang$  may enforce termination in
% a less stringent manner using available termination checkers like~\cite{approve}.

### 2017-01-23 src/Translation.tex
First version

\section{Type-Directed Translation to System F}
\label{sec:trans}

\begin{comment}

\figtwocol{f:trans}{Type-directed Translation to System F}{
\small
\bda{llrl} 
\text{Type Environments} & \tenv & ::= & \epsilon \mid \tenv,
\relation{x}{\rulet} \\
\text{Singleton Environments} & \Theta & ::= & \rulet \leadsto E\\
\text{Translation Environments} & \env & ::= & \epsilon \mid \env, \rulet \leadsto x \\
\eda \\
\bda{lc}

& \multicolumn{1}{c}{
  \myruleform{\tenv \mid \denv \turns \relation{e}{\rulet} \leadsto E}} \\
\\
\TrInt &
{ \tenv \mid \denv \turns \relation{n}{\tyint} \leadsto n } 
\\ \\

\TrVar &
\myirule
{ (\relation{x}{\rulet}) \in \tenv}
{ \tenv \mid \denv \turns \relation{x}{\rulet} \leadsto x
} 
\\ \\

\TrAbs &
\myirule
{ \tenv;\relation{x}{\rulet_1}\mid \denv \turns \relation{e}{\rulet_2} \leadsto E
}
{ \tenv\mid\denv \turns \relation{\lambda \relation{x}{\rulet_1}.e}{\rulet_1 \arrow \rulet_2}
  \leadsto \lambda \relation{x}{|\rulet_1|}.E } 
\\ \\

\TrApp &
\myirule
{ \tenv\mid \denv \turns \relation{e_1}{\rulet_1 \arrow \rulet_2} \leadsto E_1 \\
  \tenv\mid \denv \turns \relation{e_2}{\rulet_1} \leadsto E_2
}
{ \tenv\mid\denv \turns \relation{e_1\,e_2}{\rulet_2} \leadsto E_1\,E_2} 
\\ \\

\TrTAbs&
  \myirule { \tenv,\alpha \mid \env \turns \relation{e}{\rho} \leadsto
    E_1 }
           { \tenv\mid\env \turns \relation{\Lambda \alpha.e}{\forall
               \alpha.\rho} \leadsto \Lambda \alpha.E_1 } \quad\quad\quad
\\ \\
\TrTApp&
  \myirule { \tenv\mid \env \turns \relation{e}{\forall \alpha.\rho_2} \leadsto
    E}
           { \tenv\mid\env \turns \relation{e\,\rho_1}{\rho_2 [\rho_1 /\alpha]} \leadsto
    E~|\rho_1|} 
\\ \\
\TrIAbs&
  \myirule { \tenv \mid \env, \rho_1 \leadsto x \turns \relation{e}{\rho_2}  \leadsto
    E}
           { \tenv\mid\env \turns \relation{\ilambda \rho_1.e}{\rho_1 \iarrow \rho_2}  \leadsto
    \lambda \relation{x}{|\rulet_1|}. E}
\\ \\
\TrIApp&
  \myirule { \tenv \mid \env \turns \relation{e_1}{\rho_2 \iarrow
      \rho_1 \leadsto E_1} \\
             \tenv \mid \env \turns \relation{e_2}{\rho_2} \leadsto E_2}
           { \tenv\mid\env \turns \relation{e_1 \with e_2}{\rho_1} \leadsto
    E_1~E_2}
\\ \\
\TrQuery &
\myirule
{ \denv \vturns \rulet \leadsto E}
{ \tenv\mid\denv \turns \relation{?\rulet}{\rulet} \leadsto E
}

%%\TrRule &
%%\myirule
%%{ \rulet = \rulesch{\alpha}{\ruleset}{\type} \quad
%%  \vec{\alpha} \cap \mathit{ftv}(\tenv,\env) = \emptyset \\
%%  \tenv\mid\denv; \overline{\relation{\rulet}{x}} \turns 
%%  \relation{e}{\type} \leadsto E \quad\quad
%%  \bar{x}\text{~fresh} 
%%}
%%{ \tenv\mid\denv \turns 
%%  \relation{\ruleabs{\rulet}{e}}{\rulet} \leadsto
%%  \Abs{\vec{\alpha}}{\abs{({\relation{\vec{x}}{|\vec{\rulet}|}})}{E}}
%%} 
%%\\ \\

%%\TrInst &
%%\myirule
%%{ \tenv\mid\denv \turns \relation{e}{\rulesch{\alpha}{\rulesetvar}{\type}} \leadsto E
%%}
%%{ \tenv\mid\denv \turns 
%%  \relation
%%  {e[\vec{\type}]}
%%  {\subst{\vec{\alpha}}{\vec{\type}}(\rulesetvar \To \type)} 
%%  \leadsto E\;|\vec{\type}|
%%  
%%} 
%%\\ \\

%%\TrRApp &
%%\myirule
%%{ \tenv\mid\denv \turns \relation{e}{\ruleset \To \type} \leadsto E \\
%%  \tenv\mid\denv \turns \relation{e_i}{\rulet_i} \leadsto E_i \quad 
%%  (\forall \relation{e_i}{\rulet_i} \in \rulesetexp)
%%}
%%{ \tenv\mid\denv \turns \relation{(\ruleapp{e}{\rulesetexp})}{\type}
%%  \leadsto E~\vec{E}}
%%\\ \\

%%\multicolumn{2}{l}{\myruleform{\denv \vturns \rho \leadsto E}} \\ \\

%%\TrRes &
%%\myirule
%%{ 
%%  \denv(\type) = 
%%  \relation{\rulesetvar' \Rightarrow \type}{E} \quad\quad
%%  \bar{x}\text{~fresh} \\

%%  \forall \rulet_i \in \rulesetvar': 
%%  \left\{
%%    \begin{array}{ll}
%%      \denv \vturns \rulet_i \leadsto E_i &
%%      , \rulet_i \not \in \rulesetvar  \\
%%      E_i = x_i &
%%      , \rulet_i \in \rulesetvar
%%    \end{array}
%%  \right.
%%}
%%{ \denv \vturns \rulesch{\alpha}{\rulesetvar}{\type} \leadsto
%%  \Abs
%%  {\vec{\alpha}}
%%  {\abs
%%    {(\relation{\vec{x}}{|\vec{\rulet}|})}
%%    {(E\,\vec{E})}}
%%}

% \multicolumn{2}{c}{
%   \coherent(\denv, \type) \defeq 
%   \forall \theta. \theta \denv(\type) = (\theta \denv)(\theta \type) 
% }

%%\eda
%%\bda{llc}
%%\myruleform{\lookup{\env}{\type} = \rulet : E} &  
%%\myirule
%%{
%%  \lookup{\overline{\relation{\rulet}{x}}}{\type} = \rulet : E
%%}
%%{
%%  \lookup{(\env;\overline{\relation{\rulet}{x}})}{\type} = \rulet : E
%%} \\ \\
%% & 
%%\myirule
%%{
%%  \lookup{\overline{\relation{\rulet}{x}}}{\type} = \bot \quad 
%%  \lookup{\env}{\type} = \rulet
%%}
%%{
%%  \lookup{(\env;\overline{\relation{\rulet}{x}})}{\type} = \rulet
%%} \\ \\

%%\myruleform{\lookup{\overline{\relation{\rulet}{x}}}{\type} = \rulet : E} &
%% \myirule
%%{
%%  (\relation{\rulet}{x}) \in \overline{\relation{\rulet}{x}} \quad\quad \rulet = \ruleschr{\vec{\alpha}'}{\rulesetvar'}{\type'} \\
%%  \theta\type' = \type  \quad\quad \theta = [\vec{\alpha}' \mapsto \vec{\type} ]
%%}
%%{
%%  \lookup{\overline{\relation{\rulet}{x}}}{\type} = \theta\bar{\rulet}' \Rightarrow \type : x\,|\vec{\type}|
%%}

\eda \\
\begin{eqnarray*}
|\alpha| & = & \alpha \\
|\tyint| & = & \tyint \\
|\rulet_1 \arrow \rulet_2| & = & |\rulet_1| \arrow |\rulet_2| \\
|\forall \alpha. \rulet| & = & \forall \alpha. |\rulet| \\
|\rulet_1 \iarrow \rulet_2| & = & |\rulet_1| \arrow
|\rulet_2| %\\
%%|\type|_\rulet & = & |\type|_\type \\ 
%%|\forall \vec{\alpha}.\myset{\rulet_1, \cdots, \rulet_n} \To \type| & = & 
%%\forall \vec{\alpha}.|\rulet_1| \to \cdots \to |\rulet_n| \to |\type| \\
%|\tenv| & = & \myset{ (\relation{x}{|\rulet|})  ~|~ (\relation{x}{\rulet}) \in \tenv } \\
%|\denv| & = & 
%\myset{ (\relation{x}{|\rulet|}) ~|~ 
%  (\relation{\rulet}{x}) \in \denv }
\end{eqnarray*}
}

\figtwocol{fig:resolution2}{Deterministic Resolution Translation}{
\begin{center}
\framebox{$
\ba{c}
\myruleform{\env \vturns \rho \leadsto E}
\quad\quad\quad
\TrRTAbs \quad
  \myirule{\env, \rho_1 \leadsto x \vturns \rho_2 \leadsto E \quad\quad x~\mathit{fresh}}
          {\env \vturns \rho_1 \iarrow \rho_2 \leadsto
            \lambda\relation{x}{|\rho_1|}.E} 
\\ \\
\TrRIAbs \quad
  \myirule{\env \vturns \rho \leadsto E}
          {\env \vturns \forall \alpha. \rho \leadsto \Lambda\alpha.E} 
\quad\quad\quad
\TrRSimpl \quad
 \myirule{\env\langle\type\rangle = \Theta \quad\quad \env; \Theta
   \turns_\downarrow \type \leadsto E}
          {\env \vturns \type \leadsto E} 
\\ \\ \\
\myruleform{\env; \Theta \turns_\downarrow \type \leadsto E}
\quad\quad\quad
\TrIIAbs \quad
  \myirule{\env \vturns \rho_1 \leadsto E_2 \quad\quad \env; \rho_2
    \leadsto E_1~E_2
    \turns_\downarrow \type \leadsto E_3}
          {\env; \rho_1 \iarrow \rho_2 \leadsto E_1 \turns_\downarrow \type
            \leadsto E_3}  
\\ \\
\TrIDone \quad
  \myirule{}
          {\env; \type \leadsto E \turns_\downarrow \type \leadsto E}  
\quad\quad\quad
\TrITAbs \quad
  \myirule{\env; \rho[\rulet'/\alpha] \leadsto E_1~|\rulet'|\turns_\downarrow \type \leadsto
  E_2}
          {\env; \forall \alpha.\rho \leadsto E_1 \turns_\downarrow
            \type \leadsto E_2}  
\\ \\ \\
\myruleform{\env\langle\type\rangle = \Theta}
\quad\quad
\TrLHere \quad
  \myirule{\rulet \lhd \type}
          {(\env,\rulet \leadsto E)\langle\type\rangle = \rulet \leadsto E}  
\quad\quad
\TrLThere \quad
  \myirule{\rulet \mathop{\not\!\!\lhd} \type \quad\quad \env\langle\type\rangle = \Theta}
          {(\env,\rulet \leadsto E)\langle\type\rangle = \Theta}  
%%\\ \\
%%\myruleform{\rho\lhd\type} \\ \\
%%  \myirule{}
%%          {\type \leadsto E \lhd \type}
%%\quad\quad\quad
%%  \myirule{\rho[\rulet'/\alpha] \lhd \type}
%%          {\forall \alpha.\rho \lhd \type}
%%\quad\quad\quad
%%  \myirule{\rho' \lhd \type}
%%          {\rho'' \iarrow \rho' \lhd \type}
\ea
$
}
\end{center}
}

\end{comment}

In this section we explain the dynamic semantics of $\ourlang$ in terms
of System F's dynamic semantics, by means of a type-directed translation. 
This translation turns implicit contexts into explicit parameters and
statically resolves all queries, much like Wadler and Blott's dictionary
passing translation for type classes~\cite{adhoc}. 
The advantage of this approach is that we simultaneously provide a meaning to
well-typed $\ourlang$ programs and an effective implementation that resolves
all queries statically.

The translation follows the type system presented in Section~\ref{sec:ourlang}.
The additional machinery that is necessary (on top of the type system)
corresponds to the grayed parts of Figures~\ref{fig:type}, \ref{fig:resolution1} and \ref{fig:resolution2}.

%-------------------------------------------------------------------------------
\subsection{Type-Directed Translation}
Figure~\ref{fig:type} presents the translation rules that convert $\ourlang$
expressions into ones of System F. 
The gray parts of the figure essentially extend the type system with the necessary
information for the translation.
%%, but for readability we have omitted the
%%earlier gray-shaded conditions.

The syntax of System F is as follows: 
{\small
  \[ \begin{array}{llrl}
    \text{Types} & T & ::= & \alpha \mid T \arrow T 
    \mid \forall \alpha. T \\ 
    \text{Expressions} & E & ::=  & x \mid \lambda (x:T) . E \mid E\;E
    \mid \Lambda \alpha . E \mid E\;T 
  \end{array} \]}

With respect to the type system the type environments $\Gamma$ remain
the same. However, implicit environments $\env$ need to be extended with
evidence information for the translation, thus becoming
\emph{translation} environments:

{\small
  \[ \begin{array}{llrl}
\text{Translation Environments} & \env & ::= & \epsilon \mid \env, \rulet~\gbox{\leadsto E} \\
  \end{array} \]}

The main translation judgment, which adapts the typing judgment, is 
\begin{center}
  $\tenv \mid \denv \turns \relation{e}{\rho}~\gbox{\leadsto E}$
\end{center}
This judgment states that the translation of $\ourlang$ expression $e$ with
type $\rho$ is System~F expression $E$, with respect to type environment
$\tenv$ and translation
environment $\denv$. The translation environment $\denv$ relates each
rule type in the earlier implicit environment to a System~F variable
$x$; %\bruno{This should be System F expression, right?};
% \bruno{In general it is an expression right? Not limited to variables.}\tom{
% The implicit abstraction rule always introduces a variable and there is no
% other rule that extends the implicit environment.
% }
this
variable serves as explicit value-level evidence for the implicit rule.
Lookup
in the translation environment is defined similarly to lookup in the
implicit environment, except that the lookup now returns a pair of a rule type
and an evidence variable.

The function $|\cdot|$ takes 
$\ourlang$ types $\rho$ to System F types T: 
\begin{center}
\colorbox{light}{\parbox{0.3\textwidth}{\begin{eqnarray*}
|\alpha| & = & \alpha \\
%%|\tyint| & = & \tyint \\
|\rulet_1 \arrow \rulet_2| & = & |\rulet_1| \arrow |\rulet_2| \\
|\forall \alpha. \rulet| & = & \forall \alpha. |\rulet| \\
|\rulet_1 \iarrow \rulet_2| & = & |\rulet_1| \arrow |\rulet_2| %\\
\end{eqnarray*}
% \end{mdframed}
}}
\end{center}

%Note that an empty context is translated to \tyunit{} in order to delay the
%evaluation of the rule.l
%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
%\paragraph{Translating Empty Contexts}

% WC: I commented out the following paragraph because I thought this
% is too much
% For simplicity of presentation, Figure~\ref{f:trans} ignores the
% issue of empty contexts $\varnothing$. The simplified translation
% does not distinguish a rule with an empty context
% $\ruleabs{(\varnothing \To \type)}{e}$ from its body $e$.
% Nevertheless, in the operational semantics these two are different:
% a rule defers the evaluation of the rule body until explicit rule
% application. To properly capture this ``lazy'' evaluation strategy,
% we turn an empty context into an explicit \tyunit{} function
% parameter:

Variables, lambda abstractions and applications
are translated straightforwardly. Perhaps the only noteworthy 
rule is \TyIAbs. This rule associates the type $\rho_1$ with 
the fresh variable $x$ in the translation environment ($\denv$). 
This creates the necessary evidence that can be used by resolutions 
in the body of the rule abstraction to construct System F terms of type $|\rho_1|$. 
% The resolution process 
% will lookup the translation environment for a suitably matching rule. 
% If a suitable rule is found, the associated variable will then 
% be used to construct the evidence term.

\paragraph{Resolution}
The more interesting part of the translation happens when 
resolving queries. Queries are translated by rule $\TyQuery$ using the auxiliary
resolution judgment $\vturns$:
\begin{center}
$\bar{\alpha}; \env \vturns \rho~\gbox{\leadsto E}$
\end{center}

\noindent which is shown, in deterministic form, in Figure~\ref{fig:resolution2}.
%%
%% defined by rule $\TrRes$\bruno{Need to add names 
%%to the figures!}. 
The translation of resolution basically extends the type-checking
process with simultaneously building System F evidence terms. The mechanism that builds
evidence dualizes the process of peeling off abstractions and universal 
quantifiers. The rule \mylabel{R-IAbs}~wraps a lambda binder with a fresh variable 
$x$ around a System F expression $E$, which is generated from the resolution 
for the head of the rule ($\rho_2$). The rule \mylabel{R-TAbs}~wraps a type lambda binder 
around the System F expression resulting from the resolution of $\rho$.
For simple types \mylabel{R-Simp}, evidence $E_1$ is retrieved from matching 
$\type$ and then used in the resolution process of the simple type 
$\type$ with rule $\turns_\downarrow$:

\begin{center}
$\bar{\alpha};\env; \rho~\gbox{\leadsto E_1} \turns_\downarrow \type~\gbox{\leadsto E_2}$
\end{center}

The rules \mylabel{I-IAbs}~and \mylabel{I-TAbs}~are the most interesting in the translation. 
In rule \mylabel{I-IAbs}~we need evidence for $\rho_2$ to build evidence for $\type$, 
but all we have is evidence $E_1$ for $\rho_1 \iarrow \rho_2$. However, we can 
construct evidence for $\rho_2$, by generating evidence $E_2$ for $\rho_1$
and then simply applying $E_1$ to $E_2$. Rule \mylabel{I-TAbs}~is similar: we need 
evidence for $\rho$, but all we have is evidence $E_1$ for $\forall \alpha. \rho$. 
To create evidence for $\rho$ we can substitute $\alpha$ by some type 
$\rho'$ in $\rho$ and generate a term which is the type application of 
$E_1$ to $|\rho'|$.

Finally, matching ($\elookup{\env}{\type}$) also needs to be
extended with evidence generation, but this extension is
straightforward.

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
% Note how empty type contexts are also treated specially in the translation rules.
% They are translated respectively to $\tyunit$ abstractions and $\unit$ applications.

% \paragraph{Coherence of Translation}

% The type system of Section~\ref{s:types} imposes the condition
% $\welldefined(\env)$ to ensure coherence. Because translation performs
% resolution statically, it requires a stronger condition, $\stable{(\Delta)}$.
% Consider the following query:
% \begin{center}
%   $\myset{\relation{\forall \alpha. \alpha \to \alpha}{x}};
%   \myset{\relation{\tyint \to \tyint}{y}} \vturns \qask(\beta \to
%   \beta)$
% \end{center}
% where $\beta$ is a free type variable. By design, the operational semantics
% always chooses the most specific rule in the environment at runtime. For the
% sake of coherence, the \welldefined predicate in the type system
% guarantees that all possible runtime instantiations of $\beta\to\beta$ can be
% resolved.

% In contrast, the translation must decide statically on a single rule
% that is the most specific one for all possible instantiations.
% In the example, this decision cannot be made statically without knowing more about $\beta$. 
% The rule $\forall \alpha. \alpha\to\alpha$ covers all possible cases, but when
% $\beta$ is $\tyint$, it is not the most specific one. Hence, static resolution
% cannot be performed in a coherent manner. 
% The \stable{} predicate enforces static coherence, and is responsible for
% rejecting this problematic example.

% \paragraph{Incompleteness of the Translation}
% As the example above shows, not all well-typed $\ourlang$ programs
% have a translation; i.e. \\
% $\exists \denv. \welldefined(\dom(\denv)) \wedge \neg
% \stable(\denv)$. It is because $\ourlang$ does not respect
% parametricity, whereas System F does.

% From this remark on parametricity we can conclude that the operational
% semantics is a viable choice for systems where parametricity is not a
% concern and run-time type information is available. Mainstream
% object-oriented languages like Java and C\# fit this category.  On the
% other hand the elaboration semantics is a better match for functional
% programming languages concerned about parametricity (such as Haskell).

% As a remark we note that the elaboration semantics respects
% \emph{parametricity}~\cite{}, whereas the operational semantics does
% not. The example |f1| is a good representative of this fact. In
% System~F the only function that can inhabit the type |forall b . b ->
% b| is the identity function. As such, the selection of the most
% general instance in the two examples is the only possible choice for
% an elaboration into System~F. However, as we have just argued, this
% choice would be incoherent.

% The availability of run-time type information in the operational
% semantics, allows resolution to depend on that information. This 
% means that resolution can be more powerful and yet retain 
% coherence, but it also means that we lose parametricity. 
% Indeed, despite having the type |forall b . b -> b|, when applied 
% to an integer, the function |f1| is the successor (and not the
% identity) function.

%\paragraph{Correctness of Translation}

% The correctness of the translation is captured in two properties: type
% preservation and semantics preservation.
% Type preservation relates the type systems of $\ourlang$ and
% System~F. The translated expression of a well typed $\ourlang$
% expression is well typed.

\thmtranstypreserve
\begin{proof} (Sketch)
  We first prove\footnote{in the technical report} the more general lemma ``if $\tenv \mid \denv \turns
  \relation{e}{\rho} \leadsto E$, then $|\tenv|,|\denv| \turns
  \relation{E}{|\rho|}$'' by induction on the derivation of
  translation. Then, the theorem trivially follows.
\end{proof}

An important lemma in the theorem's proof is the type preservation of 
resolution.
\begin{lemma}[Type-Preserving Resolution]
Let $\env$ be an implicit environment, $\rho$ be a type and E be a System F expression.
If $\bar{\alpha};\env \vdash_r \rho \leadsto E$, then $|\env| \vdash E : |\rho|$.
\end{lemma}
% Both the theorem and the lemma are proven in Appendix~\ref{proof:preservation}.

Moreover, we can express three key properties of Figure~\ref{fig:resolution2}'s
definition of resolution in terms of the generated evidence.
\begin{lemma}[Determinacy]
The generated evidence of resolution is uniquely determined.
\[\forall \env, \rho, E_1, E_2, \bar{\alpha}: \quad\quad \bar{\alpha};\env \vdash_r \rho \leadsto E_1 ~\wedge~ \bar{\alpha};\env \vdash_r \rho \leadsto E_2 \quad\Rightarrow\quad E_1 = E_2 \]
\end{lemma}
\begin{lemma}[Soundness]
Figure~\ref{fig:resolution2}'s definition of resolution (here denoted $\vdash_r^3$)
is sound (but incomplete) with respect to Figure~\ref{fig:resolution1}'s definition
(here denoted $\vdash_r^2$).
\[\forall \env, \rho, E: \quad\quad \env \vdash_r^3 \rho \leadsto E \quad\Rightarrow\quad \env \vdash_r^2 \rho \leadsto E \]
\end{lemma}
\begin{lemma}[Coherence]
Resolution is stable under substitution.
\[\forall \env, \rho, E, \bar{\alpha}, \theta: \quad\quad \bar{\alpha};\env \vdash_r \rho \leadsto E ~\wedge~ \mathit{dom}(\theta) \cap \bar{\alpha} = \emptyset \quad\Rightarrow\quad \bar{\alpha};\theta(\env) \vdash_r \theta(\rho) \leadsto \theta(E) \]
\end{lemma}

%-------------------------------------------------------------------------------
\subsection{Evidence Generation in the Algorithm}

The evidence generation in Figure~\ref{fig:algorithm} is largely similar to
that in the deterministic specification of resolution in
Figure~\ref{fig:resolution2}.

The main difference, and complication, is due to the fact that the evidence for
type instantiation and recursive resolution is needed before these operations
actually take place, as the algorithm has to postpone them. For this reason, the
algorithm first produces placeholders that are later substituted for the actual
evidence.

The central relation is $\rho; \bar{\rho}; \bar{\alpha} \gbox{; \bar{\omega};
E}\turns_{\mathit{match}} \tau \hookrightarrow \bar{\rho}' \gbox{;
\bar{\omega}'; E'}$. It captures the matching instantiation of context type
$\rho$ against simple type $\tau$.  The input evidence for $\rho$ is $E$, and
the output evidence for the instantiation is $E'$. The accumulating parameters
$\bar{\alpha}$ and $\bar{\rho}$ denote that the instantiation of type variables
$\bar{\alpha}$ and the recursive resolution of $\bar{\rho}$ have been
postponed. We use the $\bar{\alpha}$ themselves as convenient placeholders for
the instantiating types, and we use the synthetic $\bar{\omega}$ as placeholders for
the evidence of the $\bar{\rho}$. The rules \mylabel{MTC-Abs} and \mylabel{MTC-Abs} 
introduce these two kinds of placeholders in the evidence. The former kind, $\bar{\alpha}$,
are substituted in rule \mylabel{MTC-Simp} where the actual type instantiatons $\bar{\theta}$
are computed. The latter kind, $\bar{\omega}$, are substituted later in rule \mylabel{Alg-Simp} where
the recursive resolutions take place.

Now we can state the correctness of the algorithm.
\begin{theorem}[Partial Correctness]
Let $\env$ be an implicit environment, $\rho$ be a type, E be a System F expression and $\bar{\alpha}$ a set of type variables.
Assume that $\epsilon \vdash_{\mathit{unamb}} \rho$ and also $\forall \rho_i \in \env: \epsilon \vdash_{\mathit{unamb}} \rho_i$.
Then $\bar{\alpha};\env \vdash_r \rho \leadsto E$ if and only if $\bar{\alpha};\env \vdash_{\mathit{alg}} \rho \leadsto E$,
provided that the algorithm terminates.
\end{theorem}

%-------------------------------------------------------------------------------
\subsection{Dynamic Semantics}
Finally, we define the dynamic semantics of $\ourlang$ as the composition
of the type-directed translation and System F's dynamic semantics. 
Following Siek's notation~\cite{systemfg}, this dynamic semantics is:
\[ \mathit{eval}(e) = V \quad\quad \textit{where } \epsilon\mid\epsilon \turns \relation{e}{\rho} \leadsto E \textit{ and } E \rightarrow^* V  \]
with $\rightarrow^*$ the reflexive, transitive closure of System F's standard single-step call-by-value reduction relation (see \cite[Chapter 23]{tapl}).

Now we can state the conventional type safety theorem for $\ourlang$:
\begin{theorem}[Type Safety]
If $\epsilon \mid \epsilon \turns \relation{e}{\rho}$, then $\mathit{eval}(e) = V$ for
some System F value $V$.
\end{theorem}
The proof follows trivially from Theorem~\ref{thm:type:preservation}.

### 2017-01-23 code/SourceElaborationMain.hs

@@ -1,0 +1,66 @@
+module SourceElaborationMain where
+
+import Data.Maybe (fromJust)
+
+import SrcParser
+import SourceElaboration
+import SyntaxSource
+import CoreElaboration
+import SystemFSemantics
+import System.Environment
+import qualified Data.Map as M
+
+pgm = "(? : forall a b c. {a -> a, b -> b, {a -> b} => b -> c} => a -> b -> c)"
+
+pgm2 = "((\\x : Int. x) : Int -> Int)"
+
+pgm3 = "let u : forall a. a -> a = \\x : a. x in u : Int -> Int"
+
+pgm4 = "let u : forall a. a -> a = \\x : a. x in implicit { u } in ?(Int -> Int)"
+
+pgm5 = "interface Ord a = { ord : a -> a -> Int; eq : a -> a -> Int } " ++
+       "let ordIntImpl : Ord Int = Ord { ord = ? (Int -> Int -> Int), " ++
+       "eq = ? (Int -> Int -> Int) } : Ord Int in " ++
+       "(ord : Ord Int -> Int -> Int -> Int) (ordIntImpl : Ord Int)"
+
+pgm6 = "interface Succ a = { succ : a -> a } " ++
+       "let succIntImpl : Succ Int = Succ { succ = ? (Int -> Int) } : Succ Int in " ++
+       "(succ : Succ Int -> Int -> Int) (succIntImpl : Succ Int)"
+
+pgm7 = "interface Succ a = { succ : a -> a } " ++
+       "let sc : forall a. {Succ a} => a -> a = (succ : Succ a -> a -> a) (? (Succ a)) in " ++
+       "let succIntImpl : Succ Int = Succ { succ = \\x : Int. x } : Succ Int in " ++
+       "implicit { succIntImpl } in " ++
+       "(sc : Int -> Int) 1"
+
+pgm8 = "interface Succ a = { succ : a -> a } " ++
+       "let succIntImpl : Succ Int = Succ { succ = \\x : Int. x } : Succ Int in " ++
+       "implicit { succIntImpl } in (? (Succ Int))"
+
+main = do
+  args <- getArgs
+  let pgm3' = fromJust $ srcparse pgm3
+  let pgm4' = fromJust $ srcparse pgm4
+  let pgm5' = fromJust $ srcparse pgm5
[…]

### 2017-01-24 src/Translation.tex
updating Sections 3 - 4: work in progress

@@
 \label{sec:trans}
-
-\begin{comment}
-
-\figtwocol{f:trans}{Type-directed Translation to System F}{
-\small
-\bda{llrl} 
-\text{Type Environments} & \tenv & ::= & \epsilon \mid \tenv,
-\relation{x}{\rulet} \\
-\text{Singleton Environments} & \Theta & ::= & \rulet \leadsto E\\
-\text{Translation Environments} & \env & ::= & \epsilon \mid \env, \rulet \leadsto x \\
-\eda \\
-\bda{lc}
-
-& \multicolumn{1}{c}{
-  \myruleform{\tenv \mid \denv \turns \relation{e}{\rulet} \leadsto E}} \\
-\\
-\TrInt &
-{ \tenv \mid \denv \turns \relation{n}{\tyint} \leadsto n } 
-\\ \\
-
-\TrVar &
-\myirule
-{ (\relation{x}{\rulet}) \in \tenv}
-{ \tenv \mid \denv \turns \relation{x}{\rulet} \leadsto x
-} 
-\\ \\
-
-\TrAbs &
-\myirule
-{ \tenv;\relation{x}{\rulet_1}\mid \denv \turns \relation{e}{\rulet_2} \leadsto E
-}
-{ \tenv\mid\denv \turns \relation{\lambda \relation{x}{\rulet_1}.e}{\rulet_1 \arrow \rulet_2}
-  \leadsto \lambda \relation{x}{|\rulet_1|}.E } 
-\\ \\
-
-\TrApp &
-\myirule
-{ \tenv\mid \denv \turns \relation{e_1}{\rulet_1 \arrow \rulet_2} \leadsto E_1 \\
-  \tenv\mid \denv \turns \relation{e_2}{\rulet_1} \leadsto E_2
-}
-{ \tenv\mid\denv \turns \relation{e_1\,e_2}{\rulet_2} \leadsto E_1\,E_2} 
-\\ \\
-
-\TrTAbs&
-  \myirule { \tenv,\alpha \mid \env \turns \relation{e}{\rho} \leadsto
-    E_1 }
-           { \tenv\mid\env \turns \relation{\Lambda \alpha.e}{\forall
-               \alpha.\rho} \leadsto \Lambda \alpha.E_1 } \quad\quad\quad
-\\ \\
-\TrTApp&
-  \myirule { \tenv\mid \env \turns \relation{e}{\forall \alpha.\rho_2} \leadsto
-    E}
-           { \tenv\mid\env \turns \relation{e\,\rho_1}{\rho_2 [\rho_1 /\alpha]} \leadsto
-    E~|\rho_1|} 
-\\ \\
-\TrIAbs&
-  \myirule { \tenv \mid \env, \rho_1 \leadsto x \turns \relation{e}{\rho_2}  \leadsto
-    E}
-           { \tenv\mid\env \turns \relation{\ilambda \rho_1.e}{\rho_1 \iarrow \rho_2}  \leadsto
-    \lambda \relation{x}{|\rulet_1|}. E}
-\\ \\
-\TrIApp&
-  \myirule { \tenv \mid \env \turns \relation{e_1}{\rho_2 \iarrow
-      \rho_1 \leadsto E_1} \\
-             \tenv \mid \env \turns \relation{e_2}{\rho_2} \leadsto E_2}
-           { \tenv\mid\env \turns \relation{e_1 \with e_2}{\rho_1} \leadsto
-    E_1~E_2}
-\\ \\
-\TrQuery &
-\myirule
-{ \denv \vturns \rulet \leadsto E}
-{ \tenv\mid\denv \turns \relation{?\rulet}{\rulet} \leadsto E
-}
-
-%%\TrRule &
-%%\myirule
-%%{ \rulet = \rulesch{\alpha}{\ruleset}{\type} \quad
-%%  \vec{\alpha} \cap \mathit{ftv}(\tenv,\env) = \emptyset \\
-%%  \tenv\mid\denv; \overline{\relation{\rulet}{x}} \turns 
-%%  \relation{e}{\type} \leadsto E \quad\quad
-%%  \bar{x}\text{~fresh} 
-%%}
-%%{ \tenv\mid\denv \turns 
-%%  \relation{\ruleabs{\rulet}{e}}{\rulet} \leadsto
-%%  \Abs{\vec{\alpha}}{\abs{({\relation{\vec{x}}{|\vec{\rulet}|}})}{E}}
-%%} 
-%%\\ \\
-
-%%\TrInst &
-%%\myirule
-%%{ \tenv\mid\denv \turns \relation{e}{\rulesch{\alpha}{\rulesetvar}{\type}} \leadsto E
-%%}
-%%{ \tenv\mid\denv \turns 
-%%  \relation
-%%  {e[\vec{\type}]}
-%%  {\subst{\vec{\alpha}}{\vec{\type}}(\rulesetvar \To \type)} 
-%%  \leadsto E\;|\vec{\type}|
-%%  
-%%} 
-%%\\ \\
-
-%%\TrRApp &
-%%\myirule
-%%{ \tenv\mid\denv \turns \relation{e}{\ruleset \To \type} \leadsto E \\
-%%  \tenv\mid\denv \turns \relation{e_i}{\rulet_i} \leadsto E_i \quad 
-%%  (\forall \relation{e_i}{\rulet_i} \in \rulesetexp)
-%%}
-%%{ \tenv\mid\denv \turns \relation{(\ruleapp{e}{\rulesetexp})}{\type}
-%%  \leadsto E~\vec{E}}
-%%\\ \\
-
-%%\multicolumn{2}{l}{\myruleform{\denv \vturns \rho \leadsto E}} \\ \\
-
-%%\TrRes &
-%%\myirule
-%%{ 
-%%  \denv(\type) = 
-%%  \relation{\rulesetvar' \Rightarrow \type}{E} \quad\quad
-%%  \bar{x}\text{~fresh} \\
-
-%%  \forall \rulet_i \in \rulesetvar': 
-%%  \left\{
-%%    \begin{array}{ll}
-%%      \denv \vturns \rulet_i \leadsto E_i &
-%%      , \rulet_i \not \in \rulesetvar  \\
-%%      E_i = x_i &
-%%      , \rulet_i \in \rulesetvar
-%%    \end{array}
-%%  \right.
-%%}
-%%{ \denv \vturns \rulesch{\alpha}{\rulesetvar}{\type} \leadsto
-%%  \Abs
-%%  {\vec{\alpha}}
-%%  {\abs
-%%    {(\relation{\vec{x}}{|\vec{\rulet}|})}
-%%    {(E\,\vec{E})}}
-%%}
-
-% \multicolumn{2}{c}{
-%   \coherent(\denv, \type) \defeq 
-%   \forall \theta. \theta \denv(\type) = (\theta \denv)(\theta \type) 
-% }
-
-%%\eda
-%%\bda{llc}
-%%\myruleform{\lookup{\env}{\type} = \rulet : E} &  
-%%\myirule
-%%{
-%%  \lookup{\overline{\relation{\rulet}{x}}}{\type} = \rulet : E
-%%}
-%%{
-%%  \lookup{(\env;\overline{\relation{\rulet}{x}})}{\type} = \rulet : E
-%%} \\ \\
-%% & 
-%%\myirule
-%%{
-%%  \lookup{\overline{\relation{\rulet}{x}}}{\type} = \bot \quad 
-%%  \lookup{\env}{\type} = \rulet
-%%}
-%%{
-%%  \lookup{(\env;\overline{\relation{\rulet}{x}})}{\type} = \rulet
-%%} \\ \\
-
-%%\myruleform{\lookup{\overline{\relation{\rulet}{x}}}{\type} = \rulet : E} &
-%% \myirule
-%%{
-%%  (\relation{\rulet}{x}) \in \overline{\relation{\rulet}{x}} \quad\quad \rulet = \ruleschr{\vec{\alpha}'}{\rulesetvar'}{\type'} \\
-%%  \theta\type' = \type  \quad\quad \theta = [\vec{\alpha}' \mapsto \vec{\type} ]
-%%}
-%%{
-%%  \lookup{\overline{\relation{\rulet}{x}}}{\type} = \theta\bar{\rulet}' \Rightarrow \type : x\,|\vec{\type}|
-%%}
-
-\eda \\
-\begin{eqnarray*}
-|\alpha| & = & \alpha \\
-|\tyint| & = & \tyint \\
-|\rulet_1 \arrow \rulet_2| & = & |\rulet_1| \arrow |\rulet_2| \\
-|\forall \alpha. \rulet| & = & \forall \alpha. |\rulet| \\
-|\rulet_1 \iarrow \rulet_2| & = & |\rulet_1| \arrow
-|\rulet_2| %\\
-%%|\type|_\rulet & = & |\type|_\type \\ 
-%%|\forall \vec{\alpha}.\myset{\rulet_1, \cdots, \rulet_n} \To \type| & = & 
-%%\forall \vec{\alpha}.|\rulet_1| \to \cdots \to |\rulet_n| \to |\type| \\
-%|\tenv| & = & \myset{ (\relation{x}{|\rulet|})  ~|~ (\relation{x}{\rulet}) \in \tenv } \\
-%|\denv| & = & 
-%\myset{ (\relation{x}{|\rulet|}) ~|~ 
-%  (\relation{\rulet}{x}) \in \denv }
-\end{eqnarray*}
-}
-
-\figtwocol{fig:resolution2}{Deterministic Resolution Translation}{
-\begin{center}
-\framebox{$
-\ba{c}
-\myruleform{\env \vturns \rho \leadsto E}
-\quad\quad\quad
-\TrRTAbs \quad
-  \myirule{\env, \rho_1 \leadsto x \vturns \rho_2 \leadsto E \quad\quad x~\mathit{fresh}}
-          {\env \vturns \rho_1 \iarrow \rho_2 \leadsto
-            \lambda\relation{x}{|\rho_1|}.E} 
-\\ \\
-\TrRIAbs \quad
-  \myirule{\env \vturns \rho \leadsto E}
-          {\env \vturns \forall \alpha. \rho \leadsto \Lambda\alpha.E} 
-\quad\quad\quad
-\TrRSimpl \quad
- \myirule{\env\langle\type\rangle = \Theta \quad\quad \env; \Theta
-   \turns_\downarrow \type \leadsto E}
-          {\env \vturns \type \leadsto E} 
-\\ \\ \\
-\myruleform{\env; \Theta \turns_\downarrow \type \leadsto E}
-\quad\quad\quad
-\TrIIAbs \quad
-  \myirule{\env \vturns \rho_1 \leadsto E_2 \quad\quad \env; \rho_2
-    \leadsto E_1~E_2
-    \turns_\downarrow \type \leadsto E_3}
-          {\env; \rho_1 \iarrow \rho_2 \leadsto E_1 \turns_\downarrow \type
-            \leadsto E_3}  
-\\ \\
-\TrIDone \quad
-  \myirule{}
-          {\env; \type \leadsto E \turns_\downarrow \type \leadsto E}  
-\quad\quad\quad
-\TrITAbs \quad
-  \myirule{\env; \rho[\rulet'/\alpha] \leadsto E_1~|\rulet'|\turns_\downarrow \type \leadsto
-  E_2}
-          {\env; \forall \alpha.\rho \leadsto E_1 \turns_\downarrow
-            \type \leadsto E_2}  
-\\ \\ \\
-\myruleform{\env\langle\type\rangle = \Theta}
-\quad\quad
-\TrLHere \quad
-  \myirule{\rulet \lhd \type}
-          {(\env,\rulet \leadsto E)\langle\type\rangle = \rulet \leadsto E}  
-\quad\quad
-\TrLThere \quad
-  \myirule{\rulet \mathop{\not\!\!\lhd} \type \quad\quad \env\langle\type\rangle = \Theta}
-          {(\env,\rulet \leadsto E)\langle\type\rangle = \Theta}  
-%%\\ \\
-%%\myruleform{\rho\lhd\type} \\ \\
-%%  \myirule{}
-%%          {\type \leadsto E \lhd \type}
-%%\quad\quad\quad
-%%  \myirule{\rho[\rulet'/\alpha] \lhd \type}
-%%          {\forall \alpha.\rho \lhd \type}
-%%\quad\quad\quad
-%%  \myirule{\rho' \lhd \type}
-%%          {\rho'' \iarrow \rho' \lhd \type}
-\ea
-$
-}
-\end{center}
-}
-
-\end{comment}
 
@@
 expressions into ones of System F. 
-The gray parts of the figure essentially extend the type system with the necessary
+The gray parts of the figure extend the type system with the necessary
 information for the translation.
-%%, but for readability we have omitted the
-%%earlier gray-shaded conditions.
 
@@
 
-With respect to the type system the type environments $\Gamma$ remain
-the same. However, implicit environments $\env$ need to be extended with
-evidence information for the translation, thus becoming
-\emph{translation} environments:
+The gray extension to the syntax of type environments annotates very
+implicit rule type with explicit System F evidence in the form of a 
+term variable $x$.
 
-{\small
-  \[ \begin{array}{llrl}
-\text{Translation Environments} & \env & ::= & \epsilon \mid \env, \rulet~\gbox{\leadsto E} \\
-  \end{array} \]}
-
-The main translation judgment, which adapts the typing judgment, is 
-\begin{center}
-  $\tenv \mid \denv \turns \relation{e}{\rho}~\gbox{\leadsto E}$
-\end{center}
-This judgment states that the translation of $\ourlang$ expression $e$ with
-type $\rho$ is System~F expression $E$, with respect to type environment
-$\tenv$ and translation
-environment $\denv$. The translation environment $\denv$ relates each
-rule type in the earlier implicit environment to a System~F variable
-$x$; %\bruno{This should be System F expression, right?};
-% \bruno{In general it is an expression right? Not limited to variables.}\tom{
-% The implicit abstraction rule always introduces a variable and there is no
-% other rule that extends the implicit environment.
-% }
-this
-variable serves as explicit value-level evidence for the implicit rule.
-Lookup
-in the translation environment is defined similarly to lookup in the
-implicit environment, except that the lookup now returns a pair of a rule type
-and an evidence variable.
+%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
+\paragraph{Translation of Types}
 
 The function $|\cdot|$ takes 
-$\ourlang$ types $\rho$ to System F types T: 
-\begin{center}
-\colorbox{light}{\parbox{0.3\textwidth}{\begin{eqnarray*}
+$\ourlang$ types $\rulet$ to System F types T: 
+\begin{eqnarray*}
 |\alpha| & = & \alpha \\
-%%|\tyint| & = & \tyint \\
 |\rulet_1 \arrow \rulet_2| & = & |\rulet_1| \arrow |\rulet_2| \\
@@
 \end{eqnarray*}
-% \end{mdframed}
-}}
+Its reveals that implicit $\ourlang$ arrows are translated to explicit System F function arrows.
+
+%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
+\paragraph{Translation of Terms}
+
+The type-directed translation judgment, which extends the typing judgment, is 
+\begin{center}
+  $\tenv \turns \relation{e}{\rulet}~\gbox{\leadsto E}$
 \end{center}
-
-%Note that an empty context is translated to \tyunit{} in order to delay the
-%evaluation of the rule.l
-%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
-%\paragraph{Translating Empty Contexts}
-
-% WC: I commented out the following paragraph because I thought this
-% is too much
-% For simplicity of presentation, Figure~\ref{f:trans} ignores the
-% issue of empty contexts $\varnothing$. The simplified translation
-% does not distinguish a rule with an empty context
-% $\ruleabs{(\varnothing \To \type)}{e}$ from its body $e$.
-% Nevertheless, in the operational semantics these two are different:
-% a rule defers the evaluation of the rule body until explicit rule
-% application. To properly capture this ``lazy'' evaluation strategy,
-% we turn an empty context into an explicit \tyunit{} function
-% parameter:
+This judgment states that the translation of $\ourlang$ expression $e$ with
+type $\rulet$ is System~F expression $E$, with respect to type environment
+$\tenv$.
 
@@
 are translated straightforwardly. Perhaps the only noteworthy 
-rule is \TyIAbs. This rule associates the type $\rho_1$ with 
-the fresh variable $x$ in the translation environment ($\denv$). 
+rule is \TyIAbs. This rule associates the type $\rulet_1$ with 
+the fresh variable $x$ in the type environment. 
 This creates the necessary evidence that can be used by resolutions 
-in the body of the rule abstraction to construct System F terms of type $|\rho_1|$. 
-% The resolution process 
-% will lookup the translation environment for a suitably matching rule. 
-% If a suitable rule is found, the associated variable will then 
-% be used to construct the evidence term.
+in the body of the rule abstraction to construct System F terms of type $|\rulet_1|$.
 
+%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
 \paragraph{Resolution}
-The more interesting part of the translation happens when 
-resolving queries. Queries are translated by rule $\TyQuery$ using the auxiliary
-resolution judgment $\vturns$:
-\begin{center}
-$\bar{\alpha}; \env \vturns \rho~\gbox{\leadsto E}$
-\end{center}
-
-\noindent which is shown, in deterministic form, in Figure~\ref{fig:resolution2}.
-%%
-%% defined by rule $\TrRes$\bruno{Need to add names 
-%%to the figures!}. 
-The translation of resolution basically extends the type-checking
-process with simultaneously building System F evidence terms. The mechanism that builds
-evidence dualizes the process of peeling off abstractions and universal 
+The more interesting part of the translation happens when resolving queries.
+Queries are translated by rule $\TyQuery$ using the auxiliary resolution
+judgment $\vturns$:
+\begin{equation*}
+\tenv \vturns \rulet~\gbox{\leadsto E}
+\end{equation*}
+which is shown, in deterministic form, in Figure~\ref{fig:resolution2}.
+The translation builds a System F term as evidence for the resolution.
+The mechanism that builds evidence dualizes the process of peeling off
+abstractions and universal 
 quantifiers. The rule \mylabel{R-IAbs}~wraps a lambda binder with a fresh variable 
 $x$ around a System F expression $E$, which is generated from the resolution 
-for the head of the rule ($\rho_2$). The rule \mylabel{R-TAbs}~wraps a type lambda binder 
-around the System F expression resulting from the resolution of $\rho$.
+for the head of the rule ($\rulet_2$). The rule \mylabel{R-TAbs}~wraps a type lambda binder 
+around the System F expression resulting from the resolution of $\rulet$.
+
+TODO
+
 For simple types \mylabel{R-Simp}, evidence $E_1$ is retrieved from matching 
@@
 \begin{center}
-$\bar{\alpha};\env; \rho~\gbox{\leadsto E_1} \turns_\downarrow \type~\gbox{\leadsto E_2}$
+$\bar{\alpha};\env; \rulet~\gbox{\leadsto E_1} \turns_\downarrow \type~\gbox{\leadsto E_2}$
 \end{center}
@@
 The rules \mylabel{I-IAbs}~and \mylabel{I-TAbs}~are the most interesting in the translation. 
-In rule \mylabel{I-IAbs}~we need evidence for $\rho_2$ to build evidence for $\type$, 
-but all we have is evidence $E_1$ for $\rho_1 \iarrow \rho_2$. However, we can 
-construct evidence for $\rho_2$, by generating evidence $E_2$ for $\rho_1$
+In rule \mylabel{I-IAbs}~we need evidence for $\rulet_2$ to build evidence for $\type$, 
+but all we have is evidence $E_1$ for $\rulet_1 \iarrow \rulet_2$. However, we can 
+construct evidence for $\rulet_2$, by generating evidence $E_2$ for $\rulet_1$
 and then simply applying $E_1$ to $E_2$. Rule \mylabel{I-TAbs}~is similar: we need 
-evidence for $\rho$, but all we have is evidence $E_1$ for $\forall \alpha. \rho$. 
-To create evidence for $\rho$ we can substitute $\alpha$ by some type 
-$\rho'$ in $\rho$ and generate a term which is the type application of 
-$E_1$ to $|\rho'|$.
+evidence for $\rulet$, but all we have is evidence $E_1$ for $\forall \alpha. \rulet$. 
+To create evidence for $\rulet$ we can substitute $\alpha$ by some type 
+$\rulet'$ in $\rulet$ and generate a term which is the type application of 
+$E_1$ to $|\rulet'|$.
 
@@
 
-%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
-% Note how empty type contexts are also treated specially in the translation rules.
-% They are translated respectively to $\tyunit$ abstractions and $\unit$ applications.
+\paragraph{Meta-Theory} The type-directed translation of $\ourlang$ to System F exhibits a number
+of desirable properties.
 
-% \paragraph{Coherence of Translation}
-
-% The type system of Section~\ref{s:types} imposes the condition
-% $\welldefined(\env)$ to ensure coherence. Because translation performs
-% resolution statically, it requires a stronger condition, $\stable{(\Delta)}$.
-% Consider the following query:
-% \begin{center}
-%   $\myset{\relation{\forall \alpha. \alpha \to \alpha}{x}};
-%   \myset{\relation{\tyint \to \tyint}{y}} \vturns \qask(\beta \to
-%   \beta)$
-% \end{center}
-% where $\beta$ is a free type variable. By design, the operational semantics
-% always chooses the most specific rule in the environment at runtime. For the
-% sake of coherence, the \welldefined predicate in the type system
-% guarantees that all possible runtime instantiations of $\beta\to\beta$ can be
-% resolved.
-
-% In contrast, the translation must decide statically on a single rule
-% that is the most specific one for all possible instantiations.
-% In the example, this decision cannot be made statically without knowing more about $\beta$. 
-% The rule $\forall \alpha. \alpha\to\alpha$ covers all possible cases, but when
-% $\beta$ is $\tyint$, it is not the most specific one. Hence, static resolution
-% cannot be performed in a coherent manner. 
-% The \stable{} predicate enforces static coherence, and is responsible for
-% rejecting this problematic example.
-
-% \paragraph{Incompleteness of the Translation}
-% As the example above shows, not all well-typed $\ourlang$ programs
-% have a translation; i.e. \\
-% $\exists \denv. \welldefined(\dom(\denv)) \wedge \neg
-% \stable(\denv)$. It is because $\ourlang$ does not respect
-% parametricity, whereas System F does.
-
-% From this remark on parametricity we can conclude that the operational
-% semantics is a viable choice for systems where parametricity is not a
-% concern and run-time type information is available. Mainstream
-% object-oriented languages like Java and C\# fit this category.  On the
-% other hand the elaboration semantics is a better match for functional
-% programming languages concerned about parametricity (such as Haskell).
-
-% As a remark we note that the elaboration semantics respects
-% \emph{parametricity}~\cite{}, whereas the operational semantics does
-% not. The example |f1| is a good representative of this fact. In
-% System~F the only function that can inhabit the type |forall b . b ->
-% b| is the identity function. As such, the selection of the most
-% general instance in the two examples is the only possible choice for
-% an elaboration into System~F. However, as we have just argued, this
-% choice would be incoherent.
-
-% The availability of run-time type information in the operational
-% semantics, allows resolution to depend on that information. This 
-% means that resolution can be more powerful and yet retain 
-% coherence, but it also means that we lose parametricity. 
-% Indeed, despite having the type |forall b . b -> b|, when applied 
-% to an integer, the function |f1| is the successor (and not the
-% identity) function.
-
-%\paragraph{Correctness of Translation}
-
-% The correctness of the translation is captured in two properties: type
-% preservation and semantics preservation.
-% Type preservation relates the type systems of $\ourlang$ and
-% System~F. The translated expression of a well typed $\ourlang$
-% expression is well typed.
-
-\thmtranstypreserve
-\begin{proof} (Sketch)
-  We first prove\footnote{in the technical report} the more general lemma ``if $\tenv \mid \denv \turns
-  \relation{e}{\rho} \leadsto E$, then $|\tenv|,|\denv| \turns
-  \relation{E}{|\rho|}$'' by induction on the derivation of
-  translation. Then, the theorem trivially follows.
-\end{proof}
+\begin{theorem}[Type-preserving translation]\label{thm:type:preservation} Let $e$ be an $\ourlang$
+  expression, $\rulet$ be a type, $\tenv$ a type environment and $E$ be a System F expression. If
+  $\tenv \turns \relation{e}{\rulet} \leadsto E$, then $|\tenv| \turns \relation{E}{|\rulet|}$.
+\end{theorem}
+Here we define the translation of the type environment form $\ourlang$ to System F as:
+\begin{equation*}
+\begin{array}{rcl@{\hspace{2cm}}rcl}
+|\epsilon| & = & \epsilon & |\tenv,\alpha| & = & |\tenv|, \alpha \\
+|\tenv,x : \rulet| & = & |\tenv|, x : |\rulet| &
+|\tenv, \rulet \leadsto x| & = & |\tenv|, x : |\rulet|
+\end{array}
+\end{equation*}
 
@@
 \begin{lemma}[Type-Preserving Resolution]
-Let $\env$ be an implicit environment, $\rho$ be a type and E be a System F expression.
-If $\bar{\alpha};\env \vdash_r \rho \leadsto E$, then $|\env| \vdash E : |\rho|$.
+Let $\tenv$ be a type environment, $\rulet$ be a type and E be a System F expression.
+If $\tenv \vturns \rulet \leadsto E$, then $|\tenv| \vdash E : |\rulet|$.
 \end{lemma}
@@
 The generated evidence of resolution is uniquely determined.
-\[\forall \env, \rho, E_1, E_2, \bar{\alpha}: \quad\quad \bar{\alpha};\env \vdash_r \rho \leadsto E_1 ~\wedge~ \bar{\alpha};\env \vdash_r \rho \leadsto E_2 \quad\Rightarrow\quad E_1 = E_2 \]
+\[\forall \tenv, \rulet, E_1, E_2: \quad\quad \tenv \ivturns \rulet \leadsto E_1 ~\wedge~ \tenv \ivturns \rulet \leadsto E_2 \quad\Rightarrow\quad E_1 = E_2 \]
 \end{lemma}
 \begin{lemma}[Soundness]
-Figure~\ref{fig:resolution2}'s definition of resolution (here denoted $\vdash_r^3$)
-is sound (but incomplete) with respect to Figure~\ref{fig:resolution1}'s definition
-(here denoted $\vdash_r^2$).
-\[\forall \env, \rho, E: \quad\quad \env \vdash_r^3 \rho \leadsto E \quad\Rightarrow\quad \env \vdash_r^2 \rho \leadsto E \]
+Figure~\ref{fig:resolution2}'s definition of resolution is sound (but
+incomplete) with respect to Figure~\ref{fig:resolution1}'s definition.
+\[\forall \tenv, \rulet, E: \quad\quad \tenv \ivturns \rulet \leadsto E \quad\Rightarrow\quad \tenv \vturns \rulet \leadsto E \]
 \end{lemma}
@@
 Resolution is stable under substitution.
-\[\forall \env, \rho, E, \bar{\alpha}, \theta: \quad\quad \bar{\alpha};\env \vdash_r \rho \leadsto E ~\wedge~ \mathit{dom}(\theta) \cap \bar{\alpha} = \emptyset \quad\Rightarrow\quad \bar{\alpha};\theta(\env) \vdash_r \theta(\rho) \leadsto \theta(E) \]
+\[\forall \tenv,\alpha,\tenv',\sigma,\rulet, E: \quad\quad 
+\tenv,\alpha,\tenv' \ivturns \rulet \leadsto E \quad\wedge\quad \tenv \vdash \sigma
+\quad\Rightarrow\quad 
+\tenv,\tenv'[\sigma/\alpha] \ivturns \rulet[\sigma/\alpha] \leadsto E[|\sigma|/\alpha] \]
 \end{lemma}
@@
 
-The main difference, and complication, is due to the fact that the evidence for
-type instantiation and recursive resolution is needed before these operations
-actually take place, as the algorithm has to postpone them. For this reason, the
-algorithm first produces placeholders that are later substituted for the actual
-evidence.
+% The main difference, and complication, is due to the fact that the evidence for
+% type instantiation and recursive resolution is needed before these operations
+% actually take place, as the algorithm has to postpone them. For this reason, the
+% algorithm first produces placeholders that are later substituted for the actual
+% evidence.
+% 
+% The central relation is $\rulet; \bar{\rulet}; \bar{\alpha} \gbox{; \bar{\omega};
+% E}\turns_{\mathit{match}} \tau \hookrightarrow \bar{\rulet}' \gbox{;
+% \bar{\omega}'; E'}$. It captures the matching instantiation of context type
+% $\rulet$ against simple type $\tau$.  The input evidence for $\rulet$ is $E$, and
+% the output evidence for the instantiation is $E'$. The accumulating parameters
+% $\bar{\alpha}$ and $\bar{\rulet}$ denote that the instantiation of type variables
+% $\bar{\alpha}$ and the recursive resolution of $\bar{\rulet}$ have been
+% postponed. We use the $\bar{\alpha}$ themselves as convenient placeholders for
+% the instantiating types, and we use the synthetic $\bar{\omega}$ as placeholders for
+% the evidence of the $\bar{\rulet}$. The rules \mylabel{MTC-Abs} and \mylabel{MTC-Abs} 
+% introduce these two kinds of placeholders in the evidence. The former kind, $\bar{\alpha}$,
+% are substituted in rule \mylabel{MTC-Simp} where the actual type instantiatons $\bar{\theta}$
+% are computed. The latter kind, $\bar{\omega}$, are substituted later in rule \mylabel{Alg-Simp} where
+% the recursive resolutions take place.
 
-The central relation is $\rho; \bar{\rho}; \bar{\alpha} \gbox{; \bar{\omega};
-E}\turns_{\mathit{match}} \tau \hookrightarrow \bar{\rho}' \gbox{;
-\bar{\omega}'; E'}$. It captures the matching instantiation of context type
-$\rho$ against simple type $\tau$.  The input evidence for $\rho$ is $E$, and
-the output evidence for the instantiation is $E'$. The accumulating parameters
-$\bar{\alpha}$ and $\bar{\rho}$ denote that the instantiation of type variables
-$\bar{\alpha}$ and the recursive resolution of $\bar{\rho}$ have been
-postponed. We use the $\bar{\alpha}$ themselves as convenient placeholders for
-the instantiating types, and we use the synthetic $\bar{\omega}$ as placeholders for
-the evidence of the $\bar{\rho}$. The rules \mylabel{MTC-Abs} and \mylabel{MTC-Abs} 
-introduce these two kinds of placeholders in the evidence. The former kind, $\bar{\alpha}$,
-are substituted in rule \mylabel{MTC-Simp} where the actual type instantiatons $\bar{\theta}$
-are computed. The latter kind, $\bar{\omega}$, are substituted later in rule \mylabel{Alg-Simp} where
-the recursive resolutions take place.
-
-Now we can state the correctness of the algorithm.
+With the evidence we can state the correctness of the algorithm.
 \begin{theorem}[Partial Correctness]
-Let $\env$ be an implicit environment, $\rho$ be a type, E be a System F expression and $\bar{\alpha}$ a set of type variables.
-Assume that $\epsilon \vdash_{\mathit{unamb}} \rho$ and also $\forall \rho_i \in \env: \epsilon \vdash_{\mathit{unamb}} \rho_i$.
-Then $\bar{\alpha};\env \vdash_r \rho \leadsto E$ if and only if $\bar{\alpha};\env \vdash_{\mathit{alg}} \rho \leadsto E$,
+Let $\tenv$ be a type environment, $\rulet$ be a type and E be a System F expressions.
+Assume that $\epsilon \vdash_{\mathit{unamb}} \rulet$ and also $\forall \rulet_i \in \tenv: \epsilon \vdash_{\mathit{unamb}} \rulet_i$.
+Then $\tenv \ivturns \rulet \leadsto E$ if and only if $\tenv \vdash_{\mathit{alg}} \rulet \leadsto E$,
 provided that the algorithm terminates.
@@
 \subsection{Dynamic Semantics}
-Finally, we define the dynamic semantics of $\ourlang$ as the composition
-of the type-directed translation and System F's dynamic semantics. 
-Following Siek's notation~\cite{systemfg}, this dynamic semantics is:
-\[ \mathit{eval}(e) = V \quad\quad \textit{where } \epsilon\mid\epsilon \turns \relation{e}{\rho} \leadsto E \textit{ and } E \rightarrow^* V  \]
+Finally, we define the dynamic semantics of $\ourlang$ as the composition of
+the type-directed translation and System F's dynamic semantics.  Following
+Siek's notation~\cite{systemfg}, this dynamic semantics is:
+\[ \mathit{eval}(e) = V \quad\quad \textit{where } \epsilon \turns \relation{e}{\rulet} \leadsto E \textit{ and } E \rightarrow^* V  \]
 with $\rightarrow^*$ the reflexive, transitive closure of System F's standard single-step call-by-value reduction relation (see \cite[Chapter 23]{tapl}).
@@
 \begin{theorem}[Type Safety]
-If $\epsilon \mid \epsilon \turns \relation{e}{\rho}$, then $\mathit{eval}(e) = V$ for
+If $\epsilon \turns \relation{e}{\rulet}$, then $\mathit{eval}(e) = V$ for
 some System F value $V$.

### 2017-01-30 src/Translation.tex

@@
 Queries are translated by rule $\TyQuery$ using the auxiliary resolution
-judgment $\vturns$:
+judgment $\ivturns$:
 \begin{equation*}
-\tenv \vturns \rulet~\gbox{\leadsto E}
+\tenv \ivturns \rulet~\gbox{\leadsto E}
 \end{equation*}
-which is shown, in deterministic form, in Figure~\ref{fig:resolution2}.
-The translation builds a System F term as evidence for the resolution.
+which is shown, in deterministic form, in Figure~\ref{fig:resolution2}.  The
+translation builds a System F term as evidence for the resolution.
+
 The mechanism that builds evidence dualizes the process of peeling off
-abstractions and universal 
-quantifiers. The rule \mylabel{R-IAbs}~wraps a lambda binder with a fresh variable 
-$x$ around a System F expression $E$, which is generated from the resolution 
-for the head of the rule ($\rulet_2$). The rule \mylabel{R-TAbs}~wraps a type lambda binder 
-around the System F expression resulting from the resolution of $\rulet$.
+abstractions and universal quantifiers: Rule \mylabel{R-IAbs}~wraps a lambda
+binder with a fresh variable $x$ around a System F expression $E$, which is
+generated from the resolution for the head of the rule ($\rulet_2$). Similarly,
+rule \mylabel{R-TAbs}~wraps a type lambda binder around the System F expression
+resulting from the resolution of $\rulet$.
 
-TODO
+For simple types rule $\type$ rule \mylabel{R-Simp} delegates the work of
+building evidence, when a matching rule $\rulet$ type is found in the
+environment, to rule \mylabel{L-RuleMatch}. The evidence consists of two parts:
+$E$ is the evidence of matching $\type$ against $\rulet$. This match contains
+placeholders $\bar{x}$ for the contexts whose resolution is postponed by rule
+\mylabel{M-IAbs}. It falls to rule \mylabel{L-RuleMatch} to perform these
+postponed resolutions, obtain their evidence $\bar{E}$ and fill in the
+placeholders.
 
-For simple types \mylabel{R-Simp}, evidence $E_1$ is retrieved from matching 
-$\type$ and then used in the resolution process of the simple type 
-$\type$ with rule $\turns_\downarrow$:
-
-\begin{center}
-$\bar{\alpha};\env; \rulet~\gbox{\leadsto E_1} \turns_\downarrow \type~\gbox{\leadsto E_2}$
-\end{center}
-
-The rules \mylabel{I-IAbs}~and \mylabel{I-TAbs}~are the most interesting in the translation. 
-In rule \mylabel{I-IAbs}~we need evidence for $\rulet_2$ to build evidence for $\type$, 
-but all we have is evidence $E_1$ for $\rulet_1 \iarrow \rulet_2$. However, we can 
-construct evidence for $\rulet_2$, by generating evidence $E_2$ for $\rulet_1$
-and then simply applying $E_1$ to $E_2$. Rule \mylabel{I-TAbs}~is similar: we need 
-evidence for $\rulet$, but all we have is evidence $E_1$ for $\forall \alpha. \rulet$. 
-To create evidence for $\rulet$ we can substitute $\alpha$ by some type 
-$\rulet'$ in $\rulet$ and generate a term which is the type application of 
-$E_1$ to $|\rulet'|$.
-
-Finally, matching ($\elookup{\env}{\type}$) also needs to be
-extended with evidence generation, but this extension is
-straightforward.
-
+%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
 \paragraph{Meta-Theory} The type-directed translation of $\ourlang$ to System F exhibits a number

### 2017-02-09 Main.tex
Proofs
fix latex problem
acmlarge style
make bibliography work
various small fixes
local
macro for name
change title
remove comment
ACM classification

@@
-\documentclass[sigplan, anonymous, review]{acmart}
+\documentclass[acmlarge, anonymous, review]{acmart}
 %\documentclass[prodmode,acmtoplas]{acmsmall}
@@
 \usepackage{color}
+\usepackage{xspace}
 \usepackage{graphicx}
@@
 % Title portion
-\title{The Implicit Calculus: A New Foundation for Generic Programming}
+\title{\name: Deterministic and Coherent Implicits}
 \author{Ommitted for anonymous submission}
@@
 \end{abstract}
+\begin{CCSXML}
+<ccs2012>
+<concept>
+<concept_id>10003752.10010124.10010125.10010130</concept_id>
+<concept_desc>Theory of computation~Type structures</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+<concept>
+<concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
+<concept_desc>Software and its engineering~Functional languages</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+</ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Theory of computation~Type structures}
+\ccsdesc[500]{Software and its engineering~Functional languages}
 
@@
 
-\input{src/SourceLang}
+%\input{src/SourceLang}
 
-%\input{src/Related}
+\input{src/Related}
 
 \input{src/Conclusion}
-
-% Appendix
-\appendix
-\section*{APPENDIX}
-\setcounter{section}{1}
 
@@
 \bibliographystyle{ACM-Reference-Format}
-\bibliography{sigproc}
+\bibliography{papers}
 
-%\input{src/Proofs}
+% Appendix
+\appendix
+\section*{APPENDIX}
+\setcounter{section}{1}
+
+\input{src/Proofs}
 

### 2017-02-11 src/Translation.tex
more of Phil's notes addressed

@@
 
-In this section we explain the dynamic semantics of $\ourlang$ in terms
+In this section we explain the dynamic semantics of $\name$ in terms
 of System F's dynamic semantics, by means of a type-directed translation. 
@@
 The advantage of this approach is that we simultaneously provide a meaning to
-well-typed $\ourlang$ programs and an effective implementation that resolves
+well-typed $\name$ programs and an effective implementation that resolves
 all queries statically.
@@
 \subsection{Type-Directed Translation}
-Figure~\ref{fig:type} presents the translation rules that convert $\ourlang$
+Figure~\ref{fig:type} presents the translation rules that convert $\name$
 expressions into ones of System F. 
@@
 The function $|\cdot|$ takes 
-$\ourlang$ types $\rulet$ to System F types T: 
-\begin{eqnarray*}
-|\alpha| & = & \alpha \\
-|\rulet_1 \arrow \rulet_2| & = & |\rulet_1| \arrow |\rulet_2| \\
+$\name$ types $\rulet$ to System F types T: 
+\begin{equation*}
+\begin{array}{rcl@{\hspace{1cm}}rcl}
+|\alpha| & = & \alpha &
 |\forall \alpha. \rulet| & = & \forall \alpha. |\rulet| \\
+|\rulet_1 \arrow \rulet_2| & = & |\rulet_1| \arrow |\rulet_2| &
 |\rulet_1 \iarrow \rulet_2| & = & |\rulet_1| \arrow |\rulet_2| %\\
-\end{eqnarray*}
-Its reveals that implicit $\ourlang$ arrows are translated to explicit System F function arrows.
+\end{array}
+\end{equation*}
+Its reveals that implicit $\name$ arrows are translated to explicit System F function arrows.
 
@@
 \end{center}
-This judgment states that the translation of $\ourlang$ expression $e$ with
+This judgment states that the translation of $\name$ expression $e$ with
 type $\rulet$ is System~F expression $E$, with respect to type environment
@@
 %- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
-\paragraph{Meta-Theory} The type-directed translation of $\ourlang$ to System F exhibits a number
+\paragraph{Meta-Theory} The type-directed translation of $\name$ to System F exhibits a number
 of desirable properties.
 
-\begin{theorem}[Type-preserving translation]\label{thm:type:preservation} Let $e$ be an $\ourlang$
+\begin{theorem}[Type-preserving translation]\label{thm:type:preservation} Let $e$ be an $\name$
   expression, $\rulet$ be a type, $\tenv$ a type environment and $E$ be a System F expression. If
@@
 \end{theorem}
-Here we define the translation of the type environment form $\ourlang$ to System F as:
+Here we define the translation of the type environment form $\name$ to System F as:
 \begin{equation*}
@@
 \subsection{Dynamic Semantics}
-Finally, we define the dynamic semantics of $\ourlang$ as the composition of
+Finally, we define the dynamic semantics of $\name$ as the composition of
 the type-directed translation and System F's dynamic semantics.  Following
@@
 
-Now we can state the conventional type safety theorem for $\ourlang$:
+Now we can state the conventional type safety theorem for $\name$:
 \begin{theorem}[Type Safety]

### 2017-02-16 src/Translation.tex

@@
 
-The gray extension to the syntax of type environments annotates very
+The gray extension to the syntax of type environments annotates every
 implicit rule type with explicit System F evidence in the form of a 
@@
 
-For simple types rule $\type$ rule \mylabel{R-Simp} delegates the work of
+For simple types $\type$ rule \mylabel{R-Simp} delegates the work of
 building evidence, when a matching rule $\rulet$ type is found in the
@@
 \begin{theorem}[Partial Correctness]
-Let $\tenv$ be a type environment, $\rulet$ be a type and E be a System F expressions.
+Let $\tenv$ be a type environment, $\rulet$ be a type and E be a System F expression.
 Assume that $\epsilon \vdash_{\mathit{unamb}} \rulet$ and also $\forall \rulet_i \in \tenv: \epsilon \vdash_{\mathit{unamb}} \rulet_i$.

### 2017-02-27 src/Translation.tex
Section 4

@@
 definition of resolution in terms of the generated evidence.
-\begin{lemma}[Determinacy]
-The generated evidence of resolution is uniquely determined.
-\[\forall \tenv, \rulet, E_1, E_2: \quad\quad \tenv \ivturns \rulet \leadsto E_1 ~\wedge~ \tenv \ivturns \rulet \leadsto E_2 \quad\Rightarrow\quad E_1 = E_2 \]
-\end{lemma}
+Firstly, the deterministic version of resolution is a sound variation on the original ambiguous resolution.
 \begin{lemma}[Soundness]
@@
 \end{lemma}
-\begin{lemma}[Coherence]
+Secondly, the deterministic resolution guarantees a strong form of coherence:
+\begin{lemma}[Determinacy]
+The generated evidence of resolution is uniquely determined.
+\[\forall \tenv, \rulet, E_1, E_2: \quad\quad \tenv \ivturns \rulet \leadsto E_1 ~\wedge~ \tenv \ivturns \rulet \leadsto E_2 \quad\Rightarrow\quad E_1 = E_2 \]
+\end{lemma}
+Thirdly, on top of the immediate coherence of deterministic resolution, 
+an additional coherence property holds.  
+\begin{lemma}[Stability]
 Resolution is stable under substitution.
@@
 Figure~\ref{fig:resolution2}.
+With the evidence we can state the correctness of the algorithm.
 
@@
 
-With the evidence we can state the correctness of the algorithm.
 \begin{theorem}[Partial Correctness]

### 2017-02-27 src/Conclusion.tex
add conclusion

@@
 
-Our main contribution is the development of the implicit
-calculus $\ourlang$. This calculus isolates and formalizes the key
-ideas of Scala implicits and provides a simple model for language designers 
-interested in developing similar mechanisms for their own languages. 
-In addition, $\ourlang$ supports higher-order rules and partial resolution, 
-which add considerable expressiveness to the calculus.
+This paper has presented $\ourlang$, the Calculus Of CoHerent ImplicitS, a new
+calculus for implicit programming that improves upon the implicit calculus and
+strikes a good balance between flexibility and coherence. In particular,
+$\ourlang$ supports local scoping, overlapping instances, first-class
+instances, and higher-order rules, while remaining type safe, coherent and
+unambiguous.
 
-Implicits provide an interesting alternative to conventional GP 
-mechanisms like type classes or concepts. By decoupling resolution 
-from a particular type of interfaces, implicits make resolution 
-more powerful and general. Furthermore, this decoupling has other benefits too. 
-For example, by modeling concept interfaces as conventional types, those interfaces can 
-be abstracted as any other types, avoiding the issue of second class interfaces 
-that arise with type classes or concepts.
-
-Ultimately, all the expressiveness offered by $\ourlang$
-offers a wide-range of possibilities for new generic programming applications.
+In future work it would be interesting to integrate Genus's solution for the
+global coherence problem~\cite{Zhang15LFO}.
 

### 2017-02-28 src/Conclusion.tex
Conclusion

@@
 
-This paper has presented $\ourlang$, the Calculus Of CoHerent ImplicitS, a new
+This paper presented $\ourlang$, the Calculus Of CoHerent ImplicitS, a new
 calculus for implicit programming that improves upon the implicit calculus and
 strikes a good balance between flexibility and coherence. In particular,
-$\ourlang$ supports local scoping, overlapping instances, first-class
-instances, and higher-order rules, while remaining type safe, coherent and
+$\ourlang$ supports local scoping, overlapping rules, first-class
+rules, and higher-order rules, while remaining type safe, coherent and
 unambiguous.
-
-In future work it would be interesting to integrate Genus's solution for the
-global coherence problem~\cite{Zhang15LFO}.
+Interesting future work includes integrating Genus's solution for the
+instance coherence problem~\cite{Zhang15LFO} in $\ourlang$; and 
+adding more features that show up 
+in various IP mechanisms, such as \emph{associated types}~\cite{assoctypes,assoctypes2} and \emph{type
+  families}~\cite{typefunc}.
 

### 2017-04-19 response.txt
draft response

* Presentation Improvements

  Thank you for the suggestions to improve the presentation.  We will
definitely take them into account. In particular we are very happy with the
comment on making a punchline at the end of Section 2 on stability as the key
idea; we agree that this will be very helpful to better convey the key idea to
readers.

  Tracking of variables, image what happens if we don't treat them spefically
  and consider this example resolution query

  a, a -> a |-_r forall b. b -> b

* Bugs in the typing rules

  We do indeed intentionally abuse the notation (we would not call
  that bugs), but are happy to reformulate as you suggest to clarify
  the intended meaning.

  Should we say something about Show?

* The intended type of bad on p. 5 is actually bad :: a -> a, 
  and not bad :: Trans a => a -> a, because we want to force
  the resolution of Trans at the definition of bad instead of
  the call sites of bad.

* Naming the implicit parameters only works if you want to directly
  refer to them. It does not work in the case of recursive resolution,
  e.g. Int => Bool, Int |- Bool, where multiple rules have to be combined.
  In such cases the system figures out how to combine the rules; the programmer
  does not have to write it himself.

* Provide example for \beta >_\tenv  \alpha

REVIEW 3

* We will give more practical examples of implicits, but this paper is mostly
  about restrictions to obtain coherence. The two cited papers

	B. C. d. S. Oliveira, A. Moors, and M. Odersky. 2010. Type classes as objects
	and implicits. In OOPSLA. ACM, New York, NY, USA.

	B. C. d. S. Oliveira, T. Schrijvers, W. Choi, W. Lee, and K. Yi. 2012.  e
	Implicit Calculus: A New Foundation for Generic Programming. In PLDI ’12. ACM,
	New York, NY, USA.

  provide many examples to motivate implicits.

### 2017-04-20 response.txt

@@
 
-  Tracking of variables, image what happens if we don't treat them spefically
-  and consider this example resolution query
+  We would like to clarify the need for tracking the type variables referred
+  to on p.13. Consider for example resolving the type
 
-  a, a -> a |-_r forall b. b -> b
+		ρ = b.b → b
+
+  against the environment
+
+  		Γ0 = ∀c.c → c ~> id, a, a → a ~> f
+
+  i.e.
+
+	     	Γ0 ⊢ ρ
+
+  Imagine what would happen if we did not track the type variables. Firstly,
+  using rule R-TAbs we would extend the environment with b
+
+  		Γ1 = Γ0, b
+
+  and resolve the simple type
+
+	     	Γ1 ⊢ b → b
+
+  which would get stuck on a → a when considering which rule matches in Γ1. In particular,
+  L-RuleMatch does not apply because a → a is different from b → b. Also L-RuleNoMatch does
+  not apply because a → a and b → b unify under the substitution [b := a]; thus skipping 
+  rule a → a would be deemed unstable.
+
+  However, there is an error in the above reasoning. Code inlining cannot
+  result in b being instantiated to a, because b is not free in the environment 
+  Γ0, which is the environment at the point in the code where the query happens.
+  To avoid this reasoning error, we keep track of the variables that are free
+  in the environment in Γ0. These are the variables that can be substituted, while
+  the variables added to the environment by rule R-TAbs cannot be substituted.
+
+  We will integrate this example in the text.
 
@@
 
-  Should we say something about Show?
+  TODO: Should we say something about Show?
 
@@
 
-* Provide example for \beta >_\tenv  \alpha
+* TODO: Provide example for \beta >_\tenv  \alpha
 

### 2017-04-20 response.txt
explain other judgement

@@
 
-* TODO: Provide example for \beta >_\tenv  \alpha
+* The β >Γ α judgement is related to the tracking of variables and the checking of
+  stability discussed above. 
+  It performs another necessary sanity check: a substitution [α := β] (*) due to
+  inlining is only possible if β is in scope of α. This property is verified in
+  terms of the relative positions of α and β in the environment: the latter must
+  appear before the former.
 
-REVIEW 3
+  (*) This generalizes to substitutions [α := σ] where β is a free variable in σ.
 

### 2017-04-20 response.txt
Some edits

--- 2017-04-20 response.txt
@@
+We thank all reviewers for the helpful comments.
+
 * Presentation Improvements
@@
 Thank you for the suggestions to improve the presentation.  We will
-  definitely take them into account. In particular we are very happy with the
-  comment on making a punchline at the end of Section 2 on stability as the key
-  idea; we agree that this will be very helpful to better convey the key idea to
-  readers.
+definitely take them into account. In particular we are very happy
+with the comment on making a punchline at the end of Section 2 on
+stability as the key idea; we agree that this will be very helpful to
+better convey the key idea to readers.
 
-  We would like to clarify the need for tracking the type variables referred
-  to on p.13. Consider for example resolving the type
+We would like to clarify the need for tracking the type variables
+referred to on p.13. Consider for example resolving the type
 
@@
 
-  We do indeed intentionally abuse the notation (we would not call
+  We do indeed intentionally abuse the notation (although we would not call
   that bugs), but are happy to reformulate as you suggest to clarify
@@
 * Naming the implicit parameters only works if you want to directly
-  refer to them. It does not work in the case of recursive resolution,
-  e.g. Int => Bool, Int |- Bool, where multiple rules have to be combined.
+  refer to them. It does not work in the case of recursive
+  resolution. For example consider the program:
+
+  \(Int => Bool) . \Int . ?Bool
+
+  where multiple rules have to be combined to resolve ?Bool.
   In such cases the system figures out how to combine the rules; the programmer
-  does not have to write it himself.
+  does not have to write the composition himself.
 
@@
 
-* We will give more practical examples of implicits, but this paper is mostly
-  about restrictions to obtain coherence. The two cited papers
+* Our work is on an improvement to the implicit calculus. Our
+  improvement is mainly about working out the restrictions needed to
+  ensure coherence. So it does not add any new programming features
+  per se. Practical examples illustrating the uses of
+  implicits were already presented in the following two cited papers:
 
@@
 
-	B. C. d. S. Oliveira, T. Schrijvers, W. Choi, W. Lee, and K. Yi. 2012.  e
+	B. C. d. S. Oliveira, T. Schrijvers, W. Choi, W. Lee, and K. Yi. 2012.  The
 	Implicit Calculus: A New Foundation for Generic Programming. In PLDI ’12. ACM,
@@
 
-  provide many examples to motivate implicits.
+  Thus our paper focuses only on coherence and examples about
+  coherence.

### 2017-04-20 response.txt
finish example

--- 2017-04-20 response.txt
@@
 
-* TODO: Provide example for \beta >_\tenv  \alpha
+* The β >Γ α judgement is performs another necessary sanity check during unification.
+  Consider for example resolving the type
 
-REVIEW 3
+		ρ = (∀β.β → β) → Int
+
+  against the environment with a single rule
+
+  		Γ0 = ∀α.(∀β.β → α) → Int
+
+  This will lead in rule Alg-M-Simp to the unification
+
+		mgu_{Γ1,α}(∀α.(∀β.β → α) → Int,(∀β.β → β) → Int)
+
+  with
+
+  		Γ1 = Γ0, α
+
+  After several steps, we get
+
+		mgu_{Γ2,α}(α,β)
+
+  with
+
+  		Γ2 = Γ0, α, β
+
+  The result here cannot be [β := α] because β is universally quantified
+  and thus cannot be instantiated. This is why rule U-InstR should not apply;
+  the mechanism used is that mgu is indexed in the unifiable variables --   
+  only α in this example.
+
+  However, we also cannot have [α := β] because β is not in scope at the point
+  where α has to be instantiated. We see this because β occurs after α in the 
+  environment Γ2. This scope check is captured in judgement β >Γ2 α. This check
+  is the one that prevents rule U-InstL from applying and from producing an invalid
+  substitution.
 

### 2017-04-20 response.txt
Polishing

@@
 We thank all reviewers for the helpful comments.
+
+======
+Review 1
+======
 
@@
 
-		ρ = b.b → b
+		ρ = ∀b.b → b
 
@@
 
+  In the end, by tracking the free variables, we can accept the
+  resolution above, whereas without such tracking the program would be rejected.
+
   We will integrate this example in the text.
@@
   the intended meaning.
-
-  TODO: Should we say something about Show?
 
@@
 
-* The β >Γ α judgement is performs another necessary sanity check during unification.
+* The β >Γ α judgement performs a necessary sanity check during
+  unification, which is useful to prevent some invalid unifications
+  that would arise due to scoping issues.
   Consider for example resolving the type
@@
 
-		mgu_{Γ1,α}(∀α.(∀β.β → α) → Int,(∀β.β → β) → Int)
+		mgu_{Γ1,α}((∀β.β → α) → Int,(∀β.β → β) → Int)
 
@@
 
+======
+Review 3:
+======
+
 * Our work is on an improvement to the implicit calculus. Our
@@
   coherence.
+
+  Nevertheless we will try our best to make the presentation more
+  acessible by taking all the suggestions into account.

### 2017-04-20 response.txt
small improvements

@@
 
-  		Γ0 = ∀c.c → c ~> id, a, a → a ~> f
+  		Γ0 = ∀c.c → c, a, a → a
 
@@
   However, there is an error in the above reasoning. Code inlining cannot
-  result in b being instantiated to a, because b is not free in the environment 
-  Γ0, which is the environment at the point in the code where the query happens.
-  To avoid this reasoning error, we keep track of the variables that are free
-  in the environment in Γ0. These are the variables that can be substituted, while
+  result in b being instantiated to a, because b is not in scope
+  at the point in the code where the query happens (i.e., b does not appear in Γ0).  
+  To avoid this reasoning error, we keep track of the variables that appear
+  in the environment Γ0. These are the variables that can be substituted, while
   the variables added to the environment by rule R-TAbs cannot be substituted.
@@
 
-  \(Int => Bool) . \Int . ?Bool
+  \?(Int => Bool) . \?Int . ?Bool
 
@@
 
-  Nevertheless we will try our best to make the presentation more
-  acessible by taking all the suggestions into account.
+  Nevertheless we will do our best to make the presentation more
+  accessible by taking all the suggestions into account.

### 2017-06-06 Main.tex
comments on Section 3
authoryear citations
add conclusion
jfp version: first port

@@
-\documentclass[acmlarge, anonymous, review]{acmart}
+\documentclass[acmlarge, authorname]{acmart}
 %\documentclass[prodmode,acmtoplas]{acmsmall}
@@
 \newcommand{\gbox}[1]{\colorbox{light}{$\!\!#1\!\!$}}
+\newcommand{\plw}[1]{[{\color{blue}\text{PLW:}} #1]}
 
@@
 % Document starts
+\bibliographystyle{ACM-Reference-Format}
+\citestyle{acmauthoryear}
 \begin{document}
@@
 % Page heads
-\markboth{B. Oliveira et al.}{The Implicit Calculus: A New Foundation for Generic Programming}
-
 % Title portion
 \title{\name: Deterministic and Coherent Implicits}
-\author{Ommitted for anonymous submission}
+\author{Tom Schrijvers}
+\affiliation{KU Leuven}
+\author{Bruno C.d.S. Oliveira}
+\affiliation{The University of Hong Kong}
+\author{Philip Wadler}
+\affiliation{University of Edinburgh}
 
@@
 
-\begin{comment}
-\category{D.3.2}{Programming Languages}   
-                {Language Classifications}
-                [Functional Languages, Object-Oriented Languages]
-\category{F.3.3}{Logics and Meanings of Programs}   
-                {Studies of Program Constructs}
-                []
-
-\terms{Languages}
-
-\keywords{
-Implicit parameters, type classes, C++ concepts, generic programming,
-Haskell, Scala}
-
-\acmformat{Oliveira, B. C. d. S., Schrijvers, T., Choi, W., Lee, W., Yi, K., Wadler, P.
-201?. The implicit calculus: a new foundation for generic programming.}
-
-\begin{bottomstuff}
-Author's addresses: 
-B. C. d. S. Oliveira, Department of Computer Science, Hong Kong University; 
-T. Schrijvers, Department of Applied Mathematics and Computer Science, Ghent University;
-W. Choi {and} W. Lee {and} K. Yi, \ldots, Seoul National University;
-P. Wadler, \ldots, University of Edinburgh.
-\end{bottomstuff}
-\end{comment}
-
+\setcopyright{none}
 \maketitle
@@
 
+\newpage
 % Bibliography
-\bibliographystyle{ACM-Reference-Format}
 \bibliography{papers}
 
+\newpage
 % Appendix

### 2017-06-07 jfp/src/Translation.tex
Translation: fix overflow
fix typos

--- 2017-02-27 src/Translation.tex
@@
 
-\begin{theorem}[Type-preserving translation]\label{thm:type:preservation} Let $e$ be an $\name$
-  expression, $\rulet$ be a type, $\tenv$ a type environment and $E$ be a System F expression. If
+\begin{theorem}[Type-Preserving Translation]\label{thm:type:preservation} Let $e$ be a $\name$
+  expression, $\rulet$ be a type, $\tenv$ be a type environment and $E$ be a System F expression. If
   $\tenv \turns \relation{e}{\rulet} \leadsto E$, then $|\tenv| \turns \relation{E}{|\rulet|}$.
@@
 Resolution is stable under substitution.
-\[\forall \tenv,\alpha,\tenv',\sigma,\rulet, E: \quad\quad 
-\tenv,\alpha,\tenv' \ivturns \rulet \leadsto E \quad\wedge\quad \tenv \vdash \sigma
-\quad\Rightarrow\quad 
+\[\forall \tenv,\alpha,\tenv',\sigma,\rulet, E: \enskip\
+\tenv,\alpha,\tenv' \ivturns \rulet \leadsto E \enskip\wedge\enskip \tenv \vdash \sigma
+\enskip\Rightarrow\enskip 
 \tenv,\tenv'[\sigma/\alpha] \ivturns \rulet[\sigma/\alpha] \leadsto E[|\sigma|/\alpha] \]

### 2018-04-19 jfp/src/Translation.tex
adding hyperlinks

@@
 The more interesting part of the translation happens when resolving queries.
-Queries are translated by rule $\TyQuery$ using the auxiliary resolution
+Queries are translated by rule $\rref{Ty-Query}$ using the auxiliary resolution
 judgment $\ivturns$:
@@
 The mechanism that builds evidence dualizes the process of peeling off
-abstractions and universal quantifiers: Rule \mylabel{R-IAbs}~wraps a lambda
+abstractions and universal quantifiers: Rule \rref{R-IAbs}~wraps a lambda
 binder with a fresh variable $x$ around a System F expression $E$, which is
 generated from the resolution for the head of the rule ($\rulet_2$). Similarly,
-rule \mylabel{R-TAbs}~wraps a type lambda binder around the System F expression
+rule \rref{R-TAbs}~wraps a type lambda binder around the System F expression
 resulting from the resolution of $\rulet$.
 
-For simple types $\type$ rule \mylabel{R-Simp} delegates the work of
+For simple types $\type$ rule \rref{R-Simp} delegates the work of
 building evidence, when a matching rule $\rulet$ type is found in the
-environment, to rule \mylabel{L-RuleMatch}. The evidence consists of two parts:
+environment, to rule \rref{L-RuleMatch}. The evidence consists of two parts:
 $E$ is the evidence of matching $\type$ against $\rulet$. This match contains
 placeholders $\bar{x}$ for the contexts whose resolution is postponed by rule
-\mylabel{M-IAbs}. It falls to rule \mylabel{L-RuleMatch} to perform these
+\rref{M-IAbs}. It falls to rule \rref{L-RuleMatch} to perform these
 postponed resolutions, obtain their evidence $\bar{E}$ and fill in the
@@
 \begin{lemma}[Stability]
-Resolution is stable under substitution.
+Resolution is stable under type substitution.
 \[\forall \tenv,\alpha,\tenv',\sigma,\rulet, E: \enskip\
@@
 % the instantiating types, and we use the synthetic $\bar{\omega}$ as placeholders for
-% the evidence of the $\bar{\rulet}$. The rules \mylabel{MTC-Abs} and \mylabel{MTC-Abs} 
+% the evidence of the $\bar{\rulet}$. The rules \rref{MTC-Abs} and \rref{MTC-Abs} 
 % introduce these two kinds of placeholders in the evidence. The former kind, $\bar{\alpha}$,
-% are substituted in rule \mylabel{MTC-Simp} where the actual type instantiatons $\bar{\theta}$
-% are computed. The latter kind, $\bar{\omega}$, are substituted later in rule \mylabel{Alg-Simp} where
+% are substituted in rule \rref{MTC-Simp} where the actual type instantiatons $\bar{\theta}$
+% are computed. The latter kind, $\bar{\omega}$, are substituted later in rule \rref{Alg-Simp} where
 % the recursive resolutions take place.

### 2018-05-14 jfp/src/Translation.tex
even more macros used

@@
 
-The function $|\cdot|$ takes 
+The function $\ttrans$ takes 
 $\name$ types $\rulet$ to System F types T: 
@@
 \begin{array}{rcl@{\hspace{1cm}}rcl}
-|\alpha| & = & \alpha &
-|\forall \alpha. \rulet| & = & \forall \alpha. |\rulet| \\
-|\rulet_1 \arrow \rulet_2| & = & |\rulet_1| \arrow |\rulet_2| &
-|\rulet_1 \iarrow \rulet_2| & = & |\rulet_1| \arrow |\rulet_2| %\\
+\ttrans[\alpha] & = & \alpha &
+\ttrans[\forall \alpha. \rulet] & = & \forall \alpha. \ttrans[\rulet] \\
+\ttrans[\rulet_1 \arrow \rulet_2] & = & \ttrans[\rulet_1] \arrow \ttrans[\rulet_2] &
+\ttrans[\rulet_1 \iarrow \rulet_2] & = & \ttrans[\rulet_1] \arrow \ttrans[\rulet_2] %\\
 \end{array}
@@
 \begin{center}
-  $\tenv \turns \relation{e}{\rulet}~\gbox{\leadsto E}$
+  $\wte{\tenv}{e}{\rulet}{E}$
 \end{center}
@@
 are translated straightforwardly. Perhaps the only noteworthy 
-rule is \TyIAbs. This rule associates the type $\rulet_1$ with 
+rule is \rref{Ty-IAbs}. This rule associates the type $\rulet_1$ with 
 the fresh variable $x$ in the type environment. 
 This creates the necessary evidence that can be used by resolutions 
-in the body of the rule abstraction to construct System F terms of type $|\rulet_1|$.
+in the body of the rule abstraction to construct System F terms of type $\ttrans[\rulet_1]$.
 
@@
 \begin{equation*}
-\tenv \ivturns \rulet~\gbox{\leadsto E}
+\dres{\tenv}{\rulet}{E}
 \end{equation*}
@@
   expression, $\rulet$ be a type, $\tenv$ be a type environment and $E$ be a System F expression. If
-  $\tenv \turns \relation{e}{\rulet} \leadsto E$, then $|\tenv| \turns \relation{E}{|\rulet|}$.
+  $\wte{\tenv}{e}{\rulet}{E}$, then $\fwte{\etrans{\tenv}}{E}{\ttrans[\rulet]}$.
 \end{theorem}
@@
 \begin{array}{rcl@{\hspace{2cm}}rcl}
-|\epsilon| & = & \epsilon & |\tenv,\alpha| & = & |\tenv|, \alpha \\
-|\tenv,x : \rulet| & = & |\tenv|, x : |\rulet| &
-|\tenv, \rulet \leadsto x| & = & |\tenv|, x : |\rulet|
+\etrans{\epsilon} & = & \epsilon & \etrans{\tenv,\alpha} & = & \etrans{\tenv}, \alpha \\
+\etrans{\tenv,x : \rulet} & = & \etrans{\tenv}, x : \ttrans[\rulet] &
+\etrans{\tenv, \rulet \leadsto x} & = & \etrans{\tenv}, x : \ttrans[\rulet]
 \end{array}
@@
 Let $\tenv$ be a type environment, $\rulet$ be a type and E be a System F expression.
-If $\tenv \vturns \rulet \leadsto E$, then $|\tenv| \vdash E : |\rulet|$.
+If $\ares{\tenv}{\rulet}{E}$, then $\fwte{\etrans{\tenv}}{E}{\ttrans[\rulet]}$.
 \end{lemma}
@@
 incomplete) with respect to Figure~\ref{fig:resolution1}'s definition.
-\[\forall \tenv, \rulet, E: \quad\quad \tenv \ivturns \rulet \leadsto E \quad\Rightarrow\quad \tenv \vturns \rulet \leadsto E \]
+\[\forall \tenv, \rulet, E: \quad\quad \dres{\tenv}{\rulet}{E} \quad\Rightarrow\quad \ares{\tenv}{\rulet}{E} \]
 \end{lemma}
@@
 The generated evidence of resolution is uniquely determined.
-\[\forall \tenv, \rulet, E_1, E_2: \quad\quad \tenv \ivturns \rulet \leadsto E_1 ~\wedge~ \tenv \ivturns \rulet \leadsto E_2 \quad\Rightarrow\quad E_1 = E_2 \]
+\[\forall \tenv, \rulet, E_1, E_2: \quad\quad 
+     \dres{\tenv}{\rulet}{E_1} ~\wedge~ 
+     \dres{\tenv}{\rulet}{E_2} \quad\Rightarrow\quad E_1 = E_2 \]
 \end{lemma}
@@
 \[\forall \tenv,\alpha,\tenv',\sigma,\rulet, E: \enskip\
-\tenv,\alpha,\tenv' \ivturns \rulet \leadsto E \enskip\wedge\enskip \tenv \vdash \sigma
+\dres{\tenv,\alpha,\tenv'}{\rulet}{E} \enskip\wedge\enskip \tenv \vdash \sigma
 \enskip\Rightarrow\enskip 
-\tenv,\tenv'[\sigma/\alpha] \ivturns \rulet[\sigma/\alpha] \leadsto E[|\sigma|/\alpha] \]
+\dres{\tenv,\tenv'[\sigma/\alpha]}{\rulet[\sigma/\alpha]}{E[\ttrans[\sigma]/\alpha]} \]
 \end{lemma}
@@
 Let $\tenv$ be a type environment, $\rulet$ be a type and E be a System F expression.
-Assume that $\epsilon \vdash_{\mathit{unamb}} \rulet$ and also $\forall \rulet_i \in \tenv: \epsilon \vdash_{\mathit{unamb}} \rulet_i$.
-Then $\tenv \ivturns \rulet \leadsto E$ if and only if $\tenv \vdash_{\mathit{alg}} \rulet \leadsto E$,
+Assume that $\unambig{\epsilon}{\rulet}$ and also $\forall \rulet_i \in \tenv: \unambig{\epsilon}{\rulet_i}$.
+Then $\dres{\tenv}{\rulet}{E}$ if and only if $\adres{\tenv}{\rulet}{E}$,
 provided that the algorithm terminates.
@@
 Siek's notation~\cite{systemfg}, this dynamic semantics is:
-\[ \mathit{eval}(e) = V \quad\quad \textit{where } \epsilon \turns \relation{e}{\rulet} \leadsto E \textit{ and } E \rightarrow^* V  \]
-with $\rightarrow^*$ the reflexive, transitive closure of System F's standard single-step call-by-value reduction relation (see \cite[Chapter 23]{tapl}).
+\[ \mathit{eval}(e) = V \quad\quad \textit{where } \wte{\epsilon}{e}{\rulet}{E} \textit{ and } E \rightarrow^* V  \]
+with $\rightarrow^*$ the reflexive, transitive closure of System F's standard single-step call-by-value reduction relation (see Chapter 23 of \cite{tapl}).
 
@@
 \begin{theorem}[Type Safety]
-If $\epsilon \turns \relation{e}{\rulet}$, then $\mathit{eval}(e) = V$ for
+If $\wtep{\epsilon}{e}{\rulet}$, then $\mathit{eval}(e) = V$ for
 some System F value $V$.

### 2018-05-17 jfp/src/Translation.tex
address various small issued raised by review 3
additional lemmas and theorems

@@
 Figure~\ref{fig:type} presents the translation rules that convert $\name$
-expressions into ones of System F. 
+expressions into ones of System~F. 
 The gray parts of the figure extend the type system with the necessary
@@
 Firstly, the deterministic version of resolution is a sound variation on the original ambiguous resolution.
-\begin{lemma}[Soundness]
+\begin{theorem}[Soundness]
 Figure~\ref{fig:resolution2}'s definition of resolution is sound (but
@@
 \[\forall \tenv, \rulet, E: \quad\quad \dres{\tenv}{\rulet}{E} \quad\Rightarrow\quad \ares{\tenv}{\rulet}{E} \]
-\end{lemma}
-Secondly, the deterministic resolution guarantees a strong form of coherence:
-\begin{lemma}[Determinacy]
+\end{theorem}
+Secondly, the deterministic resolution guarantees is effectively deterministic:
+\begin{theorem}[Determinacy]
 The generated evidence of resolution is uniquely determined.
@@
      \dres{\tenv}{\rulet}{E_2} \quad\Rightarrow\quad E_1 = E_2 \]
-\end{lemma}
+\end{theorem}
+It follows immediately that deterministic resolution is also coherent,
+which is formulated as follows for our elaboration-based setting.
+\begin{corollary}[Coherence]
+\[\forall \tenv, \rulet, E_1, E_2: \quad\quad 
+     \dres{\tenv}{\rulet}{E_1} ~\wedge~ 
+     \dres{\tenv}{\rulet}{E_2} \quad\Rightarrow\quad 
+     \fctx{\etrans{\tenv}}{E_1}{E_2}{\ttrans[\rulet]} \]
+\end{corollary}
+Here the judgment $\fctx{\tenv}{E_1}{E_2}{T}$
+denotes \emph{contextual equivalence}, which captures that
+$E_1$ and $E_2$ behave the same in any well-typed program context. We do not give the
+formal definition here as it is not required to prove the corollary; we only
+need to know that it is an equivalence relation, and, more spefically, that
+it is reflexive, i.e., any well-typed System~F term is contextually equivalent
+to itself.
+
 Thirdly, on top of the immediate coherence of deterministic resolution, 
-an additional coherence property holds.  
-\begin{lemma}[Stability]
+an additional stability property holds.  
+\begin{lemma}[Stability of Resolution]
 Resolution is stable under type substitution.
@@
 \dres{\tenv,\tenv'[\sigma/\alpha]}{\rulet[\sigma/\alpha]}{E[\ttrans[\sigma]/\alpha]} \]
-\end{lemma}
+\end{lemma}%
+This is a key lemma to establish that static reduction of type application 
+preserves typing and elaboration.
+\begin{lemma}[Stability of Typing under Type Application Reduction]
+Static reduction of type application preserves typing.
+\[\forall \tenv,\alpha,\sigma,\rulet, E: \enskip\
+\wte{\tenv}{(\Lambda \alpha.e)\,\suty}{\rulet}{(\Lambda \alpha.E)\,\ttrans[\suty]}
+\enskip\Rightarrow\enskip 
+\wte{\tenv}{e[\suty/\alpha]}{\rulet}{E[\ttrans[\suty]/\alpha]}
+   \]
+\end{lemma}%
+This, together with another property of System~F's contextual equivalence:
+\[ \forall \tenv, \alpha, E, T, T': \enskip\
+     \fwte{\tenv}{(\Lambda \alpha.E)\,T}{T'} \enskip\Rightarrow\enskip 
+     \fctx{\tenv}{(\Lambda \alpha.E)\,T}{E[T/\alpha]}{T'}\]
+allows us to conclude the correctness of static reduction of type application.
+\begin{theorem}[Correctness of Type Application Reduction]
+If a type application and its reduced form elaborate to two System~F terms,
+those terms are contextually equivalent.
+\begin{equation*}
+\forall \tenv, \alpha, e, \suty, E1, E2 : \enskip\\
+\wte{\tenv}{(\Lambda \alpha.e)\,\suty}{\rulet}{E_1} ~\wedge~
+\wte{\tenv}{e[\suty/\alpha]}{\rulet}{E_2}
+\enskip\Rightarrow\enskip 
+\fctx{\etrans{\tenv}}{E_1}{E_2}{\ttrans[\rulet]}
+\end{equation*}
+\end{theorem}
 

### 2018-05-29 src/DesignChoices.tex
Starting my revisions

\section{Design Choices}

Before diving into the technical details, we wish to discuss and justify a number of design 
choices taken in \name. Mostly, many of these choices are motivated by the design of Haskell 
type classes or Scala implicits.

\bruno{Cite some of the papers that discuss various design options for Haskell?}

\subsection{Predicative Instantiation}

System F is an impredicative calculus, allowing the instantiation of type variables with 
arbitrary (polymorphic) types. In contrast most functional languages only allow instantiation 
of \emph{monotypes}. Cochis is a predicative calculus.

\subsection{Committed Choice}

\name employs a committed choice for resolution. The motivation for this choice is 
largely due to the startegy employed by Haskell. Since early on it was decided that Haskell 
should not use backtracking during resolution. One key motivator for this is reasoning. Another 
one is performance.

\paragraph{Overlapping Instances}

Consider the following piece of code:

\begin{code}

data IntSet = C a deriving Eq

class C a where
  m :: a -> a

instance Eq a => C [a] where

instance Ord a => C [a] where

fun :: StablePtr Int -> [StablePtr Int]
fun sp = m [sp]

\end{code}

The point here is that StablePtr supports equality but not ordering. That is there 
is a type class instance |Eq (StablePtr a)|, but no type class instance |Ord (StablePtr a)|.

The question is should the above code type-check or not? In GHC Haskell the answer is no. 
Even though for this program there is no ambiguity: the only choice is to pick the first 
type class instance, the program is nevertheless rejected.

This is because Haskell's resolution does not consider the contexts in the resolution process:
it only considers the head of a type class instance.

\section{Superclasses}

Superclasses are not supported by \name. Here \name follows the design of Scala implicits, which 
do not have a concept similar to superclasses. In Scala a similar mechanism to 
superclasses can often be achieved with OO Subtyping and class hierarchies, although there are situations 
that don't work well. \bruno{Should I really go into this?}

Superclasses look alot like overlapping instances, for example:

\begin{code}

\end{code}

But resolution in the presence of superclasses behaves differently from overlapping instances.

Superclasses have a ad-hoc treatement in GHC.

\subsection{Coherence}

We use coherence as determinism in our paper. But, generally speaking, coherence can be interpreted
more broadly.

### 2018-06-07 src/DesignChoices.tex
Starting my revisions
Local changes

@@
 
-\begin{code}
-
-data IntSet = C a deriving Eq
-
-class C a where
-  m :: a -> a
-
-instance Eq a => C [a] where
-
-instance Ord a => C [a] where
-
-fun :: StablePtr Int -> [StablePtr Int]
-fun sp = m [sp]
-
-\end{code}
+> data IntSet = C a deriving Eq
+> 
+> class C a where
+>   m :: a -> a
+>
+> instance Eq a => C [a] where
+> 
+> instance Ord a => C [a] where
+> 
+> fun :: StablePtr Int -> [StablePtr Int]
+> fun sp = m [sp]
 
@@
 
+A few concerns are:
+
+1) open vs closed world. In open world you could add more instances, so you don't know whether your choice is unique or the best
+2) reasoning: if you backtrack then programmers have to follow the backtracking reasoning, which can be complex for more involved instance declarations
+3) performance: You may pay significant penalties at compile time by backtracking
+
 \section{Superclasses}
@@
 
-\begin{code}
-
-\end{code}
+>
 
@@
 
+special treatment of superclasses:
+
+compute the closure of all given superclasses before resolution
+
+-> treat that closure like other givens
+
+ when matching, try to match the givens first, because they are non-recursive
+
+-> if none of the givens work, try global instances
+
 \subsection{Coherence}

### 2018-06-19 jfp/src/DesignChoices.tex
Section 3

--- 2018-05-29 src/DesignChoices.tex
@@
-\section{Design Choices}
+\section{Discussion}
 
-Before diving into the technical details, we wish to discuss and justify a number of design 
-choices taken in \name. Mostly, many of these choices are motivated by the design of Haskell 
+In the design of \name we had to take several design decisions. In
+this section we discuss and justify several of those design decisions.
+Mostly, many of these choices are motivated by the design of Haskell 
 type classes or Scala implicits.

### 2018-06-19 jfp/src/Translation.tex
changes

@@
 Figure~\ref{fig:type} presents the translation rules that convert $\name$
-expressions into ones of System~F. 
+expressions into System~F expressions. 
 The gray parts of the figure extend the type system with the necessary
@@
 \end{theorem}
-Secondly, the deterministic resolution guarantees is effectively deterministic:
+Secondly, the deterministic resolution is effectively deterministic:
 \begin{theorem}[Determinacy]
@@
 Here the judgment $\fctx{\tenv}{E_1}{E_2}{T}$
-denotes \emph{contextual equivalence}, which captures that
+denotes \emph{contextual equivalence}~\cite{reynolds?}, which captures that
 $E_1$ and $E_2$ behave the same in any well-typed program context. We do not give the
@@
 to itself. 
+\bruno{Should we not mention that contextual equivalence for System F
+  is well-studied, thus interested readers can consult the definition
+  in other works?}
 

### 2018-06-20 jfp/src/Translation.tex
adapt notation in elaboration section

@@
 \end{equation*}
-Its reveals that implicit $\name$ arrows are translated to explicit System F function arrows.
+It reveals that implicit $\name$ arrows are translated to explicit System F function arrows.
 
@@
 For simple types $\type$ rule \rref{R-Simp} delegates the work of
-building evidence, when a matching rule $\rulet$ type is found in the
+building evidence, when a matching implicit $\rulet$ is found in the
 environment, to rule \rref{L-RuleMatch}. The evidence consists of two parts:
@@
 \etrans{\tenv,x : \rulet} & = & \etrans{\tenv}, x : \ttrans[\rulet] &
-\etrans{\tenv, \rulet \leadsto x} & = & \etrans{\tenv}, x : \ttrans[\rulet]
+\etrans{\tenv, \envi{\rulet}{x}} & = & \etrans{\tenv}, x : \ttrans[\rulet]
 \end{array}
@@
 Assume that $\unambig{\epsilon}{\rulet}$ and also $\forall \rulet_i \in \tenv: \unambig{\epsilon}{\rulet_i}$.
-Then $\dres{\tenv}{\rulet}{E}$ if and only if $\adres{\tenv}{\rulet}{E}$,
-provided that the algorithm terminates.
+Then $\dres{\tenv}{\rulet}{E}$ if and only if $\adres{\tenv}{\rulet}{E}$.
 \end{theorem}

### 2018-06-21 jfp/src/DesignChoices.tex
Predicativity

@@
 
-System F is an impredicative calculus, allowing the instantiation of type variables with 
-arbitrary (polymorphic) types. In contrast most functional languages only allow instantiation 
-of \emph{monotypes}. Cochis is a predicative calculus.
+System F is an impredicative calculus, allowing the instantiation of
+type variables with arbitrary (polymorphic) types. In contrast \name 
+is predicative: instantiation of type
+variables only allows \emph{monotypes}. The reason to depart from
+traditional System F is due to three factors:
+
+\begin{itemize}
+
+\item {\bf Impredicative instantiation in resolution leads to additional
+    ambiguity.} As discussed in Section~\ref{}, if instantiations of
+  type variables during resolution are allowed to be impredicative,
+  then additional ambiguity is possible. It is not obvious how to
+  remove such ambiguity while retaining impredicativity. The
+  restriction to predicative instantiation removes such ambiguity.
+
+\item {\bf Impredicativity is complex when some implicit instantiation
+  is allowed.} While traditional System F (where all type
+instantiations are explicit) is simple, matters become much more
+complicated when some implicit instantiation is allowed. Indeed the
+design of System F-like calculi with implicit instantiation and/or
+some form of type-inference is much more divided in terms of design
+choices regarding (im)predicativity. One key complication has todo
+with the most-general relation for implicit
+polymorphism. The most-general relation allows systems with implicit 
+polymorphism to decide whether a type is an instance of another type. 
+For example the type |Int -> Int| is an instance of the polymorphic 
+type |forall a . a -> a|. An algorithm for determining whether a type
+is more general than another must perform implicit instantiation. 
+Unfortunatelly it is well-known that for \emph{impredicative
+  instantiation}, such most-general relation is undecidable~\cite{}. 
+However, when only predicative instantiation is allowed, then 
+the most-general relation is decidable~\cite{}. 
+Resolution in \name is closely related to the most-general relation 
+and we believe that, under impredicative instantiation, resolution is
+indeed undecidable.\bruno{This may actually be a known result. See Rouvet?}
+
+\item {\bf Predicative instantiation is not a big restriction in
+    practice.} 
+Due to the complications brought by impredicativity, many practical 
+languages with type-inference only allow predicative instantiation.
+For example, the key algorithm for type-inference currently employed 
+by the GHC Haskell compiler is predicative~\cite{}. Other languages
+with sophisticated forms of type-inference use similar
+algorithms~\cite{}. Worth noting is that the original Hindley-Milner (HM)
+system is where the predicativity restriction on polymorphic type
+systems with type-inference was firstly put into place. 
+Since \name is intended as a target for languages for languages with 
+implicit polymorphism and type-inference, which often have
+predicativity restrictions, restricting the core language to allow
+predicative instantiation only is does not lose expressive power in practice.
+
+\end{itemize}
+
+\paragraph{Alternative Design Choices} One alternative design choice
+worth mentioning for \name would be to allow impredicative
+instantiation in type applications, but still retain the predicativity
+restriction during resolution. This design would be less restrictive
+than the design of \name, and we believe that it is a reasonable
+design. We decided to also have the predicative instantiation even 
+for the explicit type applications of \name for two reasons. Firstly, 
+since \name is aimed to be a target for source languages with
+type-inference, which often have predicative restrictions anyway, then 
+there is not much to be gained by having impredicative instantiation
+in the core. Secondly, and more importantly, some of the meta-theory
+would be more involved if impredicative instantiation on type
+applications was allowed. In particular, Lemma~\ref{} would need 
+to be generalized to account for any types, rather than just
+monotypes.
+
+Allowing full impredicativity (both in type applications and
+resolution) seems more complicated. We expect that such designs 
+are possible, but necessaraly more complicated if ambiguity and
+undecidability are to be avoided. We expect that the work on
+impredicative type-inference~\cite{} is relevant, and perhaps some 
+of the design choices employed in those works would be helpful 
+in the design of a system with full impredicativity.
 

### 2018-06-21 jfp/code/Main.hs
Cochis implementation: work in progress.
So far:

- extension with int and bool types, int and bool (true,false) literals
- parsing + pretty printing
- type checking + elaboration into System F

Still to come:

- stability check
- System F type checking, as a sanity check
- System F interpreter
- (implicit ... in ...) syntactic sugar

@@ -1,0 +1,22 @@
+module Main where
+
+import Parser
+import Syntax
+import TypeCheck
+
+topLevel :: String -> IO ()
+topLevel p =
+ let t  = parseTerm p in
+ let mct = checkTerm t in
+ do print t
+    case mct of
+      Left msg 
+        -> putStrLn msg
+      Right (ct, f) 
+        -> do print ct
+              print f
+ 
+main :: IO ()
+main
+ = getLine >>= topLevel
+ 

### 2018-06-21 jfp/code/Parser.hs

@@ -1,0 +1,205 @@
+module Parser where
+
+import Syntax
+import qualified Syntax as S
+
+import Text.Parsec
+import Text.Parsec.Expr
+import Text.Parsec.String (Parser)
+import Text.Parsec.Language (emptyDef)
+import qualified Text.Parsec.Token as Token
+
+
+-- Lexer
+
+lexer :: Token.TokenParser ()
+lexer = Token.makeTokenParser style
+  where style = emptyDef {
+          Token.reservedOpNames = ["->","=>","?","\\","\\?","/\\"],
+          Token.reservedNames = ["with","forall","int","bool","true","false"],
+          Token.commentLine = "#" }
+
+natural :: Parser Integer
+natural = Token.natural lexer
+
+parens :: Parser a -> Parser a
+parens = Token.parens lexer
+
+brackets :: Parser a -> Parser a
+brackets = Token.brackets lexer
+
+reserved :: String -> Parser ()
+reserved = Token.reserved lexer
+
+reservedOp :: String -> Parser ()
+reservedOp = Token.reservedOp lexer
+
+identifier :: Parser String
+identifier = Token.identifier lexer
+
+dot :: Parser String
+dot = Token.dot lexer
+
+colon :: Parser String
+colon = Token.colon lexer
+
+-- Parser
+
+parseContextType :: String -> ContextType
+parseContextType t = 
+  case parse (allOf context_type) "stdin" t of
+    Left err -> error (show err)
+    Right ast -> ast
+
+{-
+
+  ct1 ::= forall x. ct1
+       |  ct2
+
+  ct2 ::= ct3 (ε  |  => ct1 | -> ct4)
+
+  ct3 ::= int | bool | (ct1) | x
+
+  ct4 ::= ct3 (ε | -> ct4)
+
+ -}
+
+context_type, pcontext_type :: Parser ContextType
+(context_type, pcontext_type) = (ct1, ct3)
+  where
+
+    ct1 = ct_univ <|> ct2
+
+    ct2 = do r1 <- ct3
+             (k1 r1 <|> k2 r1 <|> return r1)
+
+    ct3 = int <|> bool <|> (CT_Simp <$> ST_TVar <$> identifier) <|> parens ct1
+
+    ct4 = do r1 <- ct3
+             (k2 r1 <|> return r1)
+
[…]

### 2018-06-21 jfp/code/Syntax.hs

@@ -1,0 +1,132 @@
+module Syntax where
+
+import Text.PrettyPrint.HughesPJ
+
+-- TYPES ----------------------
+
+type TVar = String
+
+data ContextType
+ = CT_Univ TVar ContextType
+ | CT_Rule ContextType ContextType
+ | CT_Simp SimpleType
+ deriving Eq
+
+data SimpleType
+ = ST_TVar TVar
+ | ST_Fun ContextType ContextType
+ | ST_Int
+ | ST_Bool
+ deriving Eq
+
+-- Substitution
+
+substTVar :: TVar -> SimpleType -> ContextType -> ContextType
+substTVar tv st = go1
+  where
+    go1 (CT_Univ tv' ct) 
+      = CT_Univ tv' (go1 ct)
+    go1 (CT_Rule ct1 ct2)
+      = CT_Rule (go1 ct1) (go1 ct2)
+    go1 (CT_Simp st)
+      = CT_Simp (go2 st)
+
+    go2 st'@(ST_TVar tv')
+      | tv == tv'
+      = st
+      | otherwise
+      = st'
+    go2 (ST_Fun ct1 ct2)
+      = ST_Fun (go1 ct1) (go1 ct2)
+    go2 ST_Int
+      = ST_Int
+    go2 ST_Bool
+      = ST_Bool
+
+type TSubst = [(TVar,SimpleType)]
+
+substTVars :: TSubst -> ContextType -> ContextType
+substTVars s ct
+  = foldr (uncurry substTVar) ct s
+
+
+-- TERM -----------------------
+
+type Var = String
+
+data Term
+ = TM_Var Var
+ | TM_Abs Var ContextType Term
+ | TM_App Term Term
+ | TM_TAbs TVar Term
+ | TM_TApp Term SimpleType -- restricted to monotypes
+ | TM_Query ContextType
+ | TM_IAbs ContextType Term
+ | TM_IApp Term Term
+ | TM_Int Integer
+ | TM_True
+ | TM_False
+
+-- PRETTY ------------------------
+
+parens_ p q d = 
+  if p > q then parens d else d
+
+pretty :: Pretty a => a -> Doc
+pretty x = pretty_ 0 x
+
+class Pretty a where
+  pretty_ :: Int -> a -> Doc
+
+instance Pretty ContextType where
+  pretty_ p (CT_Univ tv ct)
[…]

### 2018-06-21 jfp/code/SystemF.hs

@@ -1,0 +1,160 @@
+module SystemF where
+
+import Text.PrettyPrint.HughesPJ
+
+-- TYPES ----------------------
+
+type TVar = String
+
+data Type
+  = Univ TVar Type
+  | Fun Type Type
+  | TVar TVar
+  | IntTy
+  | BoolTy
+  deriving Eq
+
+-- Substitution
+
+substTVar :: TVar -> Type -> Type -> Type
+substTVar tv t = go
+  where
+    go (Univ tv' t) 
+      = Univ tv' (go t)
+    go (Fun t1 t2)
+      = Fun (go t1) (go t2)
+    go t'@(TVar tv')
+      | tv == tv'
+      = t
+      | otherwise
+      = t'
+    go IntTy
+      = IntTy
+    go BoolTy
+      = BoolTy
+
+type TSubst = [(TVar,Type)]
+
+substTVars :: TSubst -> Type -> Type
+substTVars s t
+  = foldr (uncurry substTVar) t s
+
+
+-- TERM -----------------------
+
+type Var = String
+
+data Term
+ = Var Var
+ | Abs Var Type Term
+ | App Term Term
+ | TAbs TVar Term
+ | TApp Term Type
+ | Int Integer
+ | True
+ | False
+
+substVar :: Var -> Term -> Term -> Term
+substVar v t = go
+  where
+    go (Var v')
+     | v == v'
+     = t
+     | otherwise
+     = Var v'
+    go (Abs v' ty t')
+     | v == v'
+     = Abs v' ty t'
+     | otherwise
+     = Abs v' ty (go t') 
+    go (App t1 t2)
+     = App (go t1) (go t2)
+    go (TAbs tv t')
+     = TAbs tv (go t')
+    go (TApp t' ty)
+     = TApp (go t') ty
+    go (Int n)
+     = Int n
+    go SystemF.True 
+     = SystemF.True
+    go SystemF.False
+     = SystemF.False
+
+
+type Subst = [(Var,Term)]
+
+substVars :: Subst -> Term -> Term
+substVars s t = foldr (uncurry substVar) t s
+
+substTVarInTerm :: TVar -> Type -> Term -> Term
+substTVarInTerm tv ty = go
+  where
+    go (Var v)
+     = Var v
+    go (Abs v ty' t)
+     = Abs v (substTVar tv ty ty') (go t)
+    go (App t1 t2)
[…]

### 2018-06-21 jfp/code/TypeCheck.hs

@@ -1,0 +1,235 @@
+module TypeCheck where
+
+import Syntax
+
+import Debug.Trace
+
+import Control.Monad.Except
+import Control.Monad.Reader
+import Control.Monad.State
+import Control.Applicative
+
+import qualified SystemF as F
+
+checkTerm :: Term -> Either String (ContextType, F.Term)
+checkTerm t 
+ = runReaderT (evalStateT (typeCheck t) 0) E_Empty
+
+-- TYPE CHECKER MONAD --------------------------------
+
+type TC = StateT Int (ReaderT Env (Either String))
+
+tcAbort :: TC a
+tcAbort = lift (throwError "Abort!")
+
+tcError :: String -> TC a
+tcError msg = throwError msg
+
+tcEnv :: TC Env
+tcEnv = ask
+
+tcFresh :: TC Var
+tcFresh =
+  do n <- get
+     put (n + 1)
+     return ("_fresh_" ++ show n)
+
+-- ENVIRONMENT --------------------------------------
+
+data Env
+  = E_Empty
+  | E_Var Var ContextType Env
+  | E_TVar TVar Env
+  | E_Imp ContextType Var Env
+
+inEnv :: (Env -> Env) -> TC a -> TC a
+inEnv = local 
+
+tcLookupVar :: Var -> TC ContextType
+tcLookupVar v = 
+  do env <- tcEnv 
+     go env
+  where
+    go E_Empty 
+      = tcError ("Variable not found: " ++ v)
+    go (E_Var w r env) 
+      | w == v     = return r
+      | otherwise  = go env
+    go (E_TVar _ env)
+      = go env
+    go (E_Imp _ _ env)
+      = go env
+
+-- TYPE CHECKER --------------------------------------
+elabCT :: ContextType -> F.Type
+elabCT (CT_Univ tv ct)
+  = F.Univ tv (elabCT ct)
+elabCT (CT_Rule ct1 ct2)
+  = F.Fun (elabCT ct1) (elabCT ct2)
+elabCT (CT_Simp st)
+  = elabST st
+
+elabST :: SimpleType -> F.Type
+elabST (ST_TVar tv)
+  = F.TVar tv
+elabST (ST_Fun ct1 ct2)
+  = F.Fun (elabCT ct1) (elabCT ct2)
+elabST (ST_Int)
+  = F.IntTy
+elabST (ST_Bool)
+  = F.BoolTy
+
+elabTSubst :: TSubst -> F.TSubst
[…]

### 2018-06-22 jfp/code/Main.hs
SystemF type checker

@@ -4,19 +4,31 @@ import Parser
 import Syntax
 import TypeCheck
 
+import qualified SystemF as F
+
 topLevel :: String -> IO ()
 topLevel p =
  let t  = parseTerm p in
  let mct = checkTerm t in
- do print t
+ do section "Parsed Term" t
     case mct of
       Left msg 
         -> putStrLn msg
       Right (ct, f) 
-        -> do print ct
-              print f
+        -> do section "Checked Type" ct
+              section "System F Elaboration" f
+              case F.checkTerm f of
+                Left msg -> putStrLn msg
+                Right ty -> do section "System F Checked Type" ty
+                               section "Type Sanity Check" (elabCT ct == ty)
  
 main :: IO ()
 main
  = getLine >>= topLevel
- 
+
+section :: Show a => String -> a -> IO ()
+section title x
+  = do putStrLn "================================================================================"
+       putStrLn title
+       putStrLn "--------------------------------------------------------------------------------"
+       print x

### 2018-06-22 jfp/code/SystemF.hs

@@ -2,6 +2,10 @@ module SystemF where
 
 import Text.PrettyPrint.HughesPJ
 
+import Control.Monad.Except
+import Control.Monad.Reader
+import Control.Applicative
+
 -- TYPES ----------------------
 
 type TVar = String
@@ -158,3 +162,82 @@ instance Show Type where
 
 instance Show Term where
   show = render . pretty
+
+-- TYPE CHECKING --------------
+
+
+checkTerm :: Term -> Either String Type
+checkTerm t 
+ = runReaderT (typeCheck t) E_Empty
+
+-- TYPE CHECKER MONAD --------------------------------
+
+type TC = ReaderT Env (Either String)
+
+tcAbort :: TC a
+tcAbort = lift (throwError "Abort!")
+
+tcError :: String -> TC a
+tcError msg = throwError msg
+
+tcEnv :: TC Env
+tcEnv = ask
+
+-- ENVIRONMENT --------------------------------------
+
+data Env
+  = E_Empty
+  | E_Var Var Type Env
+  | E_TVar TVar Env
+
+inEnv :: (Env -> Env) -> TC a -> TC a
+inEnv = local 
+
+tcLookupVar :: Var -> TC Type
+tcLookupVar v = 
+  do env <- tcEnv 
+     go env
+  where
+    go E_Empty 
+      = tcError ("Variable not found: " ++ v)
+    go (E_Var w r env) 
+      | w == v     = return r
+      | otherwise  = go env
+    go (E_TVar _ env)
+      = go env
+
+-- TYPE CHECKER --------------------------------------
+
+typeCheck :: Term -> TC Type
+typeCheck (Var v)
+ = tcLookupVar v
+typeCheck (Abs v ty1 t)
+ = do ty2 <- inEnv (E_Var v ty1) (typeCheck t)
+      return (Fun ty1 ty2)
+typeCheck (App t1 t2)
+ = do ty1 <- typeCheck t1
+      case ty1 of
+        Fun ty11 ty12 
+          -> do ty2 <- typeCheck t2
+                if ty11 == ty2
+                  then return ty12
+                  else tcError ("Function argument type mismatch, expected: " ++ show ty11 ++ ", found: " ++ show ty2)
+        _ 
+          -> tcError ("Expected function type, found: " ++ show ty1)
+typeCheck (TAbs tv t)
+ = do ty <- inEnv (E_TVar tv) (typeCheck t)
+      return (Univ tv ty)
[…]

### 2018-06-22 jfp/code/SystemF.hs
System F evaluator

@@ -241,3 +241,27 @@ typeCheck SystemF.True
  = return BoolTy
 typeCheck SystemF.False 
  = return BoolTy
+
+-- Interpreter --------------
+
+eval :: Term -> Term
+eval t@(Var _)
+ = t
+eval t@(Abs _ _ _)
+ = t
+eval (App t1 t2)
+ = case eval t1 of
+     Abs v ty t1' -> eval (substVar v (eval t2) t1')
+     t1'          -> App t1' (eval t2)
+eval t@(TAbs _ _)
+ = t
+eval (TApp t ty2) 
+ = case eval t of
+     TAbs tv t' -> eval (substTVarInTerm tv ty2 t')
+     t'         -> TApp t' ty2
+eval t@(Int _)
+ = t
+eval t@(SystemF.True)
+ = t
+eval t@(SystemF.False)
+ = t

### 2018-06-22 jfp/code/Syntax.hs
perform occurs check during unification; check for unambiguous types

@@ -2,6 +2,9 @@ module Syntax where
 
 import Text.PrettyPrint.HughesPJ
 
+import Data.Set (Set)
+import qualified Data.Set as S
+
 -- TYPES ----------------------
 
 type TVar = String
@@ -18,6 +21,40 @@ data SimpleType
  | ST_Int
  | ST_Bool
  deriving Eq
+
+isMonoTypeCT :: ContextType -> Bool
+isMonoTypeCT (CT_Simp st)
+  = isMonoTypeST st
+isMonoTypeCT _
+  = False
+
+isMonoTypeST :: SimpleType -> Bool
+isMonoTypeST (ST_TVar _)
+  = True
+isMonoTypeST (ST_Fun ct1 ct2)
+  = isMonoTypeCT ct1 && isMonoTypeCT ct2
+isMonoTypeST ST_Int
+  = True
+isMonoTypeST ST_Bool
+  = True
+
+freeTVarsCT :: ContextType -> Set TVar
+freeTVarsCT (CT_Univ tv ct)
+  = S.delete tv (freeTVarsCT ct)
+freeTVarsCT (CT_Rule ct1 ct2)
+  = freeTVarsCT ct1 `S.union` freeTVarsCT ct2
+freeTVarsCT (CT_Simp st)
+  = freeTVarsST st
+
+freeTVarsST :: SimpleType -> Set TVar
+freeTVarsST (ST_TVar tv)
+  = S.singleton tv
+freeTVarsST (ST_Fun ct1 ct2)
+  = freeTVarsCT ct1 `S.union` freeTVarsCT ct2
+freeTVarsST ST_Int
+  = S.empty
+freeTVarsST ST_Bool
+  = S.empty
 
 -- Substitution
 

### 2018-06-22 jfp/code/TypeCheck.hs

@@ -11,6 +11,8 @@ import Control.Applicative
 
 import qualified SystemF as F
 
+import qualified Data.Set as S
+
 checkTerm :: Term -> Either String (ContextType, F.Term)
 checkTerm t 
  = runReaderT (evalStateT (typeCheck t) 0) E_Empty
@@ -33,6 +35,14 @@ tcFresh =
   do n <- get
      put (n + 1)
      return ("_fresh_" ++ show n)
+
+tcGuard :: Bool -> String -> TC ()
+tcGuard b msg
+  | b
+  = return ()
+  | otherwise
+  = tcError msg
+  
 
 -- ENVIRONMENT --------------------------------------
 
@@ -114,7 +124,8 @@ typeCheck (TM_TApp t st)
           -> tcError ("Expected universal type, found: " ++ show ct)
          
 typeCheck (TM_IAbs ct1 t)
- = do x <- tcFresh
+ = do tcGuard (unambigCT ct1) ("Implicit abstraction type should be unambiguous: " ++ show ct1)
+      x <- tcFresh
       (ct2, f) <- inEnv (E_Imp ct1 x) (typeCheck t)
       return (CT_Rule ct1 ct2, F.Abs x (elabCT ct1) f)
 typeCheck (TM_IApp t1 t2)
@@ -130,7 +141,9 @@ typeCheck (TM_IApp t1 t2)
           -> tcError ("Expected rule type, found: " ++ show ct1)
       
 typeCheck (TM_Query ct)
- = do f <- resolve ct
+ = do 
+      tcGuard (unambigCT ct) ("Query type is ambiguous: " ++ show ct)
+      f <- resolve ct
       return (ct, f)
 typeCheck (TM_Int n)
  = return (CT_Simp ST_Int, F.Int n)
@@ -203,6 +216,8 @@ unify st1 st2 tvars =
                return []
     go1 (ST_TVar tv1) st2@(ST_Fun ct1 ct2)
       = do guard (tv1 `elem` tvars)
+           guard (not (occursST tv1 st2))
+           guard (isMonoTypeST st2)
            return [(tv1,st2)]
     go1 (ST_Fun ct1 ct2) (ST_TVar tv2)
       = empty
@@ -233,3 +248,18 @@ unify st1 st2 tvars =
       = go1 st1 st2
     go2 _ _
       = empty 
+
+    occursST :: TVar -> SimpleType -> Bool
[…]

### 2018-06-25 jfp/code/Main.hs
add termination check to implementation

@@ -14,8 +14,9 @@ topLevel p =
     case mct of
       Left msg 
         -> putStrLn msg
-      Right (ct, f) 
-        -> do section "Checked Type" ct
+      Right ((ct, f), msgs)
+        -> do mapM_ putStrLn msgs
+              section "Checked Type" ct
               section "System F Elaboration" f
               section "System F Evaluation" (F.eval f)
               case F.checkTerm f of

### 2018-06-25 jfp/code/TypeCheck.hs

@@ -7,19 +7,20 @@ import Debug.Trace
 import Control.Monad.Except
 import Control.Monad.Reader
 import Control.Monad.State
+import Control.Monad.Writer
 import Control.Applicative
 
 import qualified SystemF as F
 
 import qualified Data.Set as S
 
-checkTerm :: Term -> Either String (ContextType, F.Term)
+checkTerm :: Term -> Either String ((ContextType, F.Term),[String])
 checkTerm t 
- = runReaderT (evalStateT (typeCheck t) 0) E_Empty
+ = runWriterT (runReaderT (evalStateT (typeCheck t) 0) E_Empty)
 
 -- TYPE CHECKER MONAD --------------------------------
 
-type TC = StateT Int (ReaderT Env (Either String))
+type TC = StateT Int (ReaderT Env (WriterT [String] (Either String)))
 
 tcAbort :: TC a
 tcAbort = lift (throwError "Abort!")
@@ -42,7 +43,14 @@ tcGuard b msg
   = return ()
   | otherwise
   = tcError msg
-  
+
+tcWarn :: Bool -> String -> TC ()
+tcWarn b msg
+  | b
+  = return ()
+  | otherwise
+  = tell ["Warning: " ++ msg]
+    
 
 -- ENVIRONMENT --------------------------------------
 
@@ -125,6 +133,7 @@ typeCheck (TM_TApp t st)
          
 typeCheck (TM_IAbs ct1 t)
  = do tcGuard (unambigCT ct1) ("Implicit abstraction type should be unambiguous: " ++ show ct1)
+      tcWarn (termCT ct1) ("Termination check violated for " ++ show ct1 ++ ". It's up to you.")
       x <- tcFresh
       (ct2, f) <- inEnv (E_Imp ct1 x) (typeCheck t)
       return (CT_Rule ct1 ct2, F.Abs x (elabCT ct1) f)
@@ -143,6 +152,7 @@ typeCheck (TM_IApp t1 t2)
 typeCheck (TM_Query ct)
  = do 
       tcGuard (unambigCT ct) ("Query type is ambiguous: " ++ show ct)
+      tcWarn (termCT ct) ("Termination check violated for " ++ show ct ++ ". It's up to you.")
       f <- resolve ct
       return (ct, f)
 typeCheck (TM_Int n)
@@ -263,3 +273,38 @@ unambigCT = go S.empty
[…]

### 2018-06-25 jfp/code/TypeCheck.hs
check ordering on type variables

@@ -308,3 +308,36 @@ termCT (CT_Rule ct1 ct2)
 
 termCT (CT_Simp st)
   = True
+
+beforeTVar :: TVar -> TVar -> TC Bool
+beforeTVar tv1 tv2 =
+  tcEnv >>= go1
+  where
+    go1 :: Env -> TC Bool
+    go1 E_Empty
+      = tcError ("Type variables not found in environment: " ++ show tv1 ++ ", " ++ show tv2)
+    go1 (E_Var _ _ env)
+      = go1 env
+    go1 (E_TVar tv env)
+      | tv == tv1
+      = go2 env
+      | tv == tv2
+      = return False
+      | otherwise
+      = go1 env
+    go1 (E_Imp _ _ env)
+      = go1 env
+
+    go2 E_Empty
+      = tcError ("Type variable not found in environment: " ++ show tv2)
+    go2 (E_Var _ _ env)
+      = go2 env
+    go2 (E_TVar tv env)
+      | tv == tv2
+      = return True
+      | otherwise
+      = go2 env
+    go2 (E_Imp _ _ env)
+      = go2 env
+    
+

### 2018-06-26 jfp/code/TypeCheck.hs
stability check

@@ -50,7 +50,10 @@ tcWarn b msg
   = return ()
   | otherwise
   = tell ["Warning: " ++ msg]
-    
+
+tcNot :: TC a -> TC ()
+tcNot p
+  =  join ((p >> return (tcError "not")) <|> (return (return ())))
 
 -- ENVIRONMENT --------------------------------------
 
@@ -77,6 +80,20 @@ tcLookupVar v =
       = go env
     go (E_Imp _ _ env)
       = go env
+
+freeTVarsEnv :: TC [TVar]
+freeTVarsEnv
+  = do env <- tcEnv
+       return (go env)
+    where
+      go E_Empty
+        = []
+      go (E_Var _ _ env)
+        = go env
+      go (E_TVar tv env)
+        = tv : go env
+      go (E_Imp _ _ env)
+        = go env
 
 -- TYPE CHECKER --------------------------------------
 elabCT :: ContextType -> F.Type
@@ -164,8 +181,13 @@ typeCheck TM_False
 
 
 resolve :: ContextType -> TC (F.Term)
-resolve ct = res ct
-  where
+resolve ct = do tvars <- freeTVarsEnv 
+                main tvars ct
+ where
+
+  main inittvars ct = res ct
+
+   where
 
     res :: ContextType -> TC (F.Term)
     res (CT_Univ tv ct)
@@ -193,10 +215,10 @@ resolve ct = res ct
                         )
              ) 
          <|> 
-             (do f <- lkp env st
+             (do tcNot (sta ct st inittvars)
+                 f <- lkp env st
                  return (return f))
         ) >>= \c -> c
-        -- TODO: test for stability
  
     mtc :: ContextType -> F.Term -> SimpleType -> [(Var,ContextType)] -> [TVar] 
         -> TC (F.Term, [(Var,ContextType)])
@@ -209,28 +231,59 @@ resolve ct = res ct
     mtc (CT_Univ tv ct) f st cts tvars
       = mtc ct (F.TApp f (F.TVar tv)) st cts (tv:tvars)
 
+    sta :: ContextType -> SimpleType -> [TVar] -> TC Bool
+    sta (CT_Simp st') st tvars
+      = do s <- unify st' st tvars
+           return True
+    sta (CT_Rule ct1 ct2) st tvars
+      = sta ct2 st tvars
+    sta (CT_Univ tv ct) st tvars
[…]

### 2018-06-26 jfp/src/Translation.tex

@@
 It follows immediately that deterministic resolution is also coherent,
-which is formulated as follows for our elaboration-based setting.
+which for our elaboration-based setting is formulated, following Biernacki and Polesiuk~\shortcite{dariusz}, in
+terms of \emph{contextual equivalence}~\cite{morris}.
 \begin{corollary}[Coherence]
@@
 \end{corollary}
-Here the judgment $\fctx{\tenv}{E_1}{E_2}{T}$
-denotes \emph{contextual equivalence}~\cite{reynolds?}, which captures that
-$E_1$ and $E_2$ behave the same in any well-typed program context. We do not give the
-formal definition here as it is not required to prove the corollary; we only
-need to know that it is an equivalence relation, and, more spefically, that
-it is reflexive, i.e., any well-typed System~F term is contextually equivalent
-to itself. 
-\bruno{Should we not mention that contextual equivalence for System F
-  is well-studied, thus interested readers can consult the definition
-  in other works?}
+Here the contextual equivalence judgment $\fctx{\tenv}{E_1}{E_2}{T}$ captures
+that $E_1$ and $E_2$ behave the same in any well-typed program context. We do
+not give the formal definition here\footnote{A definition can be found at \url{https://www.seas.harvard.edu/courses/cs252/2017fa/lectures/lec02-contextual-equiv.pdf}} as it is not required to prove the
+corollary; we only need to know that it is an equivalence relation, and, more
+spefically, that it is reflexive, i.e., any well-typed System~F term is
+contextually equivalent to itself. 
+% \bruno{Should we not mention that contextual equivalence for System F
+%   is well-studied, thus interested readers can consult the definition
+%   in other works?}
 

### 2018-06-27 jfp/code/Parser.hs
a few more small features

@@ -16,7 +16,7 @@ lexer :: Token.TokenParser ()
 lexer = Token.makeTokenParser style
   where style = emptyDef {
           Token.reservedOpNames = ["->","=>","?","\\","\\?","/\\"],
-          Token.reservedNames = ["with","forall","int","bool","true","false","implicit","in"],
+          Token.reservedNames = ["with","forall","int","bool","true","false","implicit","in","plus"],
           Token.commentLine = "#" }
 
 natural :: Parser Integer
@@ -115,13 +115,13 @@ mono_type = mt1
 
 {-
 
-   t ::= implicit t : ρ in t | λx:ρ.t | λ?ρ.t | t[σ] | t t | t with t | x | (t)
+   t ::=  implicit t : ρ in t | λx:ρ.t | λ?ρ.t | t[σ] | t t | t with t | x | (t)
 
    t0 ::= implicit t0 : ρ in t0 | Λa.t0 | ?ct3 | λ (x : ρ. t0 | ?ρ.t0 ) | 
 
         | t1 ( | [σ] | with t0 | t1) 
 
-   t1 ::= n | true | false | x | (t0)
+   t1 ::= plus | n | true | false | x | (t0)
 
  -}
 
@@ -146,6 +146,7 @@ term = t0
     t1 =     (TM_Int <$> natural)
          <|> (TM_True <$ reserved "true")
          <|> (TM_False <$ reserved "false")
+         <|> (TM_Plus <$ reserved "plus")
          <|> (TM_Var <$> identifier) 
          <|> parens t0
 

### 2018-06-27 jfp/code/TypeCheck.hs

@@ -53,7 +53,7 @@ tcWarn b msg
 
 tcNot :: TC a -> TC ()
 tcNot p
-  =  join ((p >> return (tcError "not")) <|> (return (return ())))
+  =  join ((p >> return (tcError "not stable")) <|> (return (return ())))
 
 -- ENVIRONMENT --------------------------------------
 
@@ -178,6 +178,8 @@ typeCheck TM_True
  = return (CT_Simp ST_Bool, F.True)
 typeCheck TM_False
  = return (CT_Simp ST_Bool, F.False)
+typeCheck TM_Plus
+ = return (CT_Simp (ST_Fun (CT_Simp ST_Int) (CT_Simp (ST_Fun (CT_Simp ST_Int) (CT_Simp ST_Int)))), F.Plus)
 
 
 resolve :: ContextType -> TC (F.Term)
@@ -242,7 +244,7 @@ resolve ct = do tvars <- freeTVarsEnv
 
 unify :: SimpleType -> SimpleType -> [TVar] -> TC TSubst
 unify st1 st2 tvars = 
-     trace ("unify " ++ show st1 ++ " " ++ show st2 ++ " " ++ show tvars) $
+     -- trace ("unify " ++ show st1 ++ " " ++ show st2 ++ " " ++ show tvars) $
      let ss = go1 st1 st2
      in do env <- tcEnv
            case [s | s <- ss, validTSubst env s]  of
@@ -257,9 +259,9 @@ unify st1 st2 tvars =
       
 
     go1 :: SimpleType -> SimpleType -> [TSubst]
-    go1 st1 st2
-      | trace ("go1 " ++ show st1 ++ " " ++ show st2) $ False
-      = error "UNREACHABLE"
+--     go1 st1 st2
+--       | trace ("go1 " ++ show st1 ++ " " ++ show st2) $ False
+--       = error "UNREACHABLE"
     go1 st1@(ST_TVar tv1) st2@(ST_TVar tv2)
       =    (do guard (tv1 `elem` tvars)
                guard (tv1 /= tv2)

### 2018-06-28 jfp/src/Translation.tex
real update theorem statements

@@
 \begin{theorem}[Type-Preserving Translation]\label{thm:type:preservation} Let $e$ be a $\name$
-  expression, $\rulet$ be a type, $\tenv$ be a type environment and $E$ be a System F expression. If
+  expression, $\rulet$ be a type, $\tenv$ a type environment and $E$ a System F expression. If
   $\wte{\tenv}{e}{\rulet}{E}$, then $\fwte{\etrans{\tenv}}{E}{\ttrans[\rulet]}$.
@@
 \begin{theorem}[Determinacy]
-The generated evidence of resolution is uniquely determined.
-\[\forall \tenv, \rulet, E_1, E_2: \quad\quad 
+%The generated evidence of resolution is uniquely determined.
+Let $\tenv$ be a type environment, $\rulet$ be a type and $E_1$, $E_2$ be system F expressions.
+Assume that $\unambig{}{\rulet}$ and $\forall ?\rulet_i\in\tenv$, $\unambig{}{\rulet_i}$. Then
+\[ %\forall \tenv, \rulet, E_1, E_2: \quad\quad 
      \dres{\tenv}{\rulet}{E_1} ~\wedge~ 
@@
 \begin{corollary}[Coherence]
-\[\forall \tenv, \rulet, E_1, E_2: \quad\quad 
+Let $\tenv$ be a type environment, $\rulet$ be a type and $E_1$, $E_2$ be system F expressions.
+Assume that $\unambig{}{\rulet}$ and $\forall ?\rulet_i\in\tenv$, $\unambig{}{\rulet_i}$. Then
+\[ %\forall \tenv, \rulet, E_1, E_2: \quad\quad 
      \dres{\tenv}{\rulet}{E_1} ~\wedge~ 
@@
 Resolution is stable under type substitution.
-\[\forall \tenv,\alpha,\tenv',\sigma,\rulet, E: \enskip\
+\[\forall \tenv,\alpha,\tenv',\suty,\rulet, E: \enskip\
 \dres{\tenv,\alpha,\tenv'}{\rulet}{E} \enskip\wedge\enskip \tenv \vdash \sigma
@@
 Let $\tenv$ be a type environment, $\rulet$ be a type and E be a System F expression.
-Assume that $\unambig{\epsilon}{\rulet}$ and also $\forall \rulet_i \in \tenv: \unambig{\epsilon}{\rulet_i}$.
+Assume that $\unambig{}{\rulet}$ and also $\forall ?\rulet_i \in \tenv: \unambig{}{\rulet_i}$.
 Then $\dres{\tenv}{\rulet}{E}$ if and only if $\adres{\tenv}{\rulet}{E}$.

### 2018-06-29 jfp/src/Translation.tex
small clarification

@@
 \begin{theorem}[Soundness]
-Figure~\ref{fig:resolution2}'s definition of resolution is sound (but
-incomplete) with respect to Figure~\ref{fig:resolution1}'s definition.
+Figure~\ref{fig:resolution2}'s deterministic definition of resolution is sound (but
+incomplete) with respect to Figure~\ref{fig:resolution1}'s ambiguous definition.
 \[\forall \tenv, \rulet, E: \quad\quad \dres{\tenv}{\rulet}{E} \quad\Rightarrow\quad \ares{\tenv}{\rulet}{E} \]

### 2018-07-02 jfp/src/Translation.tex
point to appendix for proofs

@@
 resolution.
-\begin{lemma}[Type-Preserving Resolution]
+\begin{lemma}[Type-Preserving Resolution]\label{lem:type:preservation}
 Let $\tenv$ be a type environment, $\rulet$ be a type and E be a System F expression.
@@
 \end{lemma}
-% Both the theorem and the lemma are proven in Appendix~\ref{proof:preservation}.
+Section~\ref{proof:preservation} of the Appendix points to the mechanized proofs
+of Theorem~\ref{thm:type:preservation} and Lemma~\ref{lem:type:preservation} stated above.
 
@@
 \end{theorem}
+Section~\ref{proof:det_soundness} of the Appendix points to the mechanized proof of the above
+theorem.
 Secondly, the deterministic resolution is effectively deterministic:
-\begin{theorem}[Determinacy]
+\begin{theorem}[Determinacy]\label{thm:determinacy}
 %The generated evidence of resolution is uniquely determined.
@@
 \end{theorem}
+A full proof of Theorem~\ref{thm:determinacy} can be found in
+Section~\ref{proof:determinism} of the Appendix.
+
 It follows immediately that deterministic resolution is also coherent,
@@
 an additional stability property holds.
-\begin{lemma}[Stability of Resolution]
+\begin{lemma}[Stability of Resolution]\label{lem:stable:resolution}
 Resolution is stable under type substitution.
@@
 \begin{lemma}[Stability of Typing under Type Application Reduction]
+\label{lem:stable:typing}
 Static reduction of type application preserves typing.
@@
 \end{theorem}
-
+The stability Lemmas~\ref{lem:stable:resolution} and~\ref{lem:stable:typing} are proved in
+Section~\ref{proof:coherence} of the Appendix.
 %-------------------------------------------------------------------------------
@@
 \end{theorem}
+The proof is split in two parts, one for each ``direction'' of the theorem.
+Section~\ref{proof:alg_soundness} proves
+soundness of the algorithm with respect to deterministic resolution, while Section~\ref{proof:alg_completeness} proves partial completeness.
 

### 2018-07-02 jfp/src/Translation.tex
Section 61

@@
 allows us to conclude the correctness of static reduction of type application.
-\begin{theorem}[Correctness of Type Application Reduction]
+\begin{theorem}[Correctness of Type Application Reduction]\label{lem:stable:correct}
 If a type application and its reduced form elaborate to two System~F terms,

### 2018-07-11 jfp/src/Translation.tex
make rule/implicit terminology consistent

@@
 The gray extension to the syntax of type environments annotates every
-implicit rule type with explicit System F evidence in the form of a 
+implicit's type with explicit System F evidence in the form of a 
 term variable $x$.
@@
 terms of \emph{contextual equivalence}~\cite{morris}.
-\begin{corollary}[Coherence]
+\begin{corollary}[Coherence]\label{lem:coherence}
 Let $\tenv$ be a type environment, $\rulet$ be a type and $E_1$, $E_2$ be system F expressions.

### 2018-07-11 jfp/src/Translation.tex
abbreviate Rule(No)Match to (No)Match

@@
 building evidence, when a matching implicit $\rulet$ is found in the
-environment, to rule \rref{L-RuleMatch}. The evidence consists of two parts:
+environment, to rule \rref{L-Match}. The evidence consists of two parts:
 $E$ is the evidence of matching $\type$ against $\rulet$. This match contains
 placeholders $\bar{x}$ for the contexts whose resolution is postponed by rule
-\rref{M-IAbs}. It falls to rule \rref{L-RuleMatch} to perform these
+\rref{M-IAbs}. It falls to rule \rref{L-Match} to perform these
 postponed resolutions, obtain their evidence $\bar{E}$ and fill in the

### 2018-07-11 jfp/src/Translation.tex
fix typos

@@
 corollary; we only need to know that it is an equivalence relation, and, more
-spefically, that it is reflexive, i.e., any well-typed System~F term is
+specifically, that it is reflexive, i.e., any well-typed System~F term is
 contextually equivalent to itself. 

### 2018-11-19 jfp/src/Translation.tex
fix hyperref
Reviewer 1, partially addressed

@@
 $\name$ types $\rulet$ to System F types T: 
-\begin{equation*}
+\begin{myequation*}
 \begin{array}{rcl@{\hspace{1cm}}rcl}
@@
 \end{array}
-\end{equation*}
+\end{myequation*}
 It reveals that implicit $\name$ arrows are translated to explicit System F function arrows.
@@
 judgment $\ivturns$:
-\begin{equation*}
+\begin{myequation*}
 \dres{\tenv}{\rulet}{E}
-\end{equation*}
+\end{myequation*}
 which is shown, in deterministic form, in Figure~\ref{fig:resolution2}.  The
@@
 placeholders $\bar{x}$ for the contexts whose resolution is postponed by rule
-\rref{M-IAbs}. It falls to rule \rref{L-Match} to perform these
+\rref{M-IApp}. It falls to rule \rref{L-Match} to perform these
 postponed resolutions, obtain their evidence $\bar{E}$ and fill in the
@@
 \end{theorem}
-Here we define the translation of the type environment form $\name$ to System F as:
-\begin{equation*}
+Here we define the translation of the type environment from $\name$ to System F as:
+\begin{myequation*}
 \begin{array}{rcl@{\hspace{2cm}}rcl}
@@
 \end{array}
-\end{equation*}
+\end{myequation*}
 
@@
 
-Moreover, we can express three key properties of Figure~\ref{fig:resolution2}'s
+In addition, we can express three key properties of Figure~\ref{fig:resolution2}'s
 definition of resolution in terms of the generated evidence.
-Firstly, the deterministic version of resolution is a sound variation on the original ambiguous resolution.
+Firstly, the deterministic version of resolution is a sound variation on the original ambiguous
+resolution.
 \begin{theorem}[Soundness]
@@
 \end{theorem}
-Section~\ref{proof:det_soundness} of the Appendix points to the mechanized proof of the above
-theorem.
+As a corollary (by use of this result with Lemma~\ref{lem:type:preservation}), the
+deterministic resolution process is also type-preserving. Section~\ref{proof:det_soundness} of
+the Appendix points to the mechanized proof of the above theorem.
 Secondly, the deterministic resolution is effectively deterministic:
@@
 \begin{lemma}[Stability of Resolution]\label{lem:stable:resolution}
-Resolution is stable under type substitution.
+Resolution is stable under monotype substitution.
 \[\forall \tenv,\alpha,\tenv',\suty,\rulet, E: \enskip\
@@
 those terms are contextually equivalent.
-\begin{equation*}
+\begin{myequation*}
 \forall \tenv, \alpha, e, \suty, E1, E2 : \enskip\\
@@
 \fctx{\etrans{\tenv}}{E_1}{E_2}{\ttrans[\rulet]}
-\end{equation*}
+\end{myequation*}
 \end{theorem}
@@
 Section~\ref{proof:alg_soundness} proves
-soundness of the algorithm with respect to deterministic resolution, while Section~\ref{proof:alg_completeness} proves partial completeness.
+soundness of the algorithm with respect to deterministic resolution, while Section~\ref{proof:alg_completeness} proves partial completeness. The latter correctness property is partial because
+it does not hold without the additional termination conditions for the queried type and the
+type environment, \(\tenv\).
 
@@
 \end{theorem}
-The proof follows trivially from Theorem~\ref{thm:type:preservation}.
+The proof follows from Theorem~\ref{thm:type:preservation} and System F's well-known
+normalization property.

### 2018-11-21 jfp/src/Translation.tex
more fixes

@@
 \end{theorem}
+Section~\ref{proof:det_soundness} of
+the Appendix points to the mechanized proof of the above theorem.
 As a corollary (by use of this result with Lemma~\ref{lem:type:preservation}), the
-deterministic resolution process is also type-preserving. Section~\ref{proof:det_soundness} of
-the Appendix points to the mechanized proof of the above theorem.
+deterministic resolution process is also type-preserving.
 Secondly, the deterministic resolution is effectively deterministic:
@@
 \dres{\tenv,\tenv'[\sigma/\alpha]}{\rulet[\sigma/\alpha]}{E[\ttrans[\sigma]/\alpha]} \]
-\end{lemma}%
+\end{lemma}
 This is a key lemma to establish that static reduction of type application 
@@
 Let $\tenv$ be a type environment, $\rulet$ be a type and E be a System F expression.
-Assume that $\unambig{}{\rulet}$ and also $\forall ?\rulet_i \in \tenv: \unambig{}{\rulet_i}$.
+Assume that $\unambig{}{\rulet}$ and also $\unambig{}{\rulet_i}$ for all \(?\rulet_i \in \tenv\).
 Then $\dres{\tenv}{\rulet}{E}$ if and only if $\adres{\tenv}{\rulet}{E}$.

### 2018-11-27 jfp/src/Translation.tex

@@
 \end{corollary}
-Here the contextual equivalence judgment $\fctx{\tenv}{E_1}{E_2}{T}$ captures
+Here, the well-studied~\cite{contextual_equivalence} contextual equivalence judgment $\fctx{\tenv}{E_1}{E_2}{T}$ captures
 that $E_1$ and $E_2$ behave the same in any well-typed program context. We do
-not give the formal definition here\footnote{A definition can be found at \url{https://www.seas.harvard.edu/courses/cs252/2017fa/lectures/lec02-contextual-equiv.pdf}} as it is not required to prove the
+not give the formal definition here as it is not required to prove the
+%\footnote{A definition can be found at \url{https://www.seas.harvard.edu/courses/cs252/2017fa/lectures/lec02-contextual-equiv.pdf}} 
 corollary; we only need to know that it is an equivalence relation, and, more
